---
title: "Tetracyclines and the risk of *Clostridioides difficile* infection"
subtitle: "Updated systematic review & Bayesian random-effects meta-analysis"
date: today
bibliography: references.bib
link-citations: true
format:
  html:
    toc: false
    theme: cosmo
    fig-align: center
    default-image-extension: png
  pdf:
    documentclass: scrartcl
    fig-align: center
    default-image-extension: pdf
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r packages}
# Packages must load in an uncached chunk: on a cache hit the setup chunk is
# not re-executed, so library() calls placed there would never run.
library(tidyverse)
library(bayesmeta)
library(metafor)
library(knitr)
```

```{r setup}
#| cache: true
es   <- read_csv("tbl_es_analysis_ready.csv", show_col_types = FALSE)
prim <- es |> filter(use_in_primary == "Yes")

# PECO comparator: another antibiotic. Studies whose only comparator is
# "no antibiotic" are held out for a separate secondary analysis.
prim_abx   <- prim |> filter(comparator_type != "no_antibiotic")   # k = 11 (primary)
prim_noabx <- prim |> filter(comparator_type == "no_antibiotic")   # k = 9  (secondary)

hn <- function(t) dhalfnormal(t, scale = 0.5)   # Half-Normal(0, 0.5) on tau

fit_prior <- function(df, mean, sd) {
  bayesmeta(y = df$yi, sigma = df$sei, labels = df$reference,
            mu.prior.mean = mean, mu.prior.sd = sd, tau.prior = hn)
}

summ <- function(f) {
  sm <- f$summary
  tibble(
    OR = exp(sm["median","mu"]), OR_lo = exp(sm["95% lower","mu"]), OR_hi = exp(sm["95% upper","mu"]),
    tau = sm["median","tau"], tau_lo = sm["95% lower","tau"], tau_hi = sm["95% upper","tau"],
    P_OR_lt_1 = f$pposterior(mu = 0), P_OR_lt_080 = f$pposterior(mu = log(0.80)),
    pred_lo = exp(sm["95% lower","theta"]), pred_hi = exp(sm["95% upper","theta"])
  )
}

# ---- Primary set (k = 11), four priors ----
fits_abx <- list(
  primary    = fit_prior(prim_abx, 0,         1),
  vague      = fit_prior(prim_abx, 0,         10),
  sceptical  = fit_prior(prim_abx, 0,         0.35),
  literature = fit_prior(prim_abx, log(0.62), 0.10)
)
res_abx <- imap_dfr(fits_abx, ~ summ(.x) |> mutate(prior = .y, .before = 1))

# ---- Secondary set: no-antibiotic comparator (k = 9) ----
sec <- summ(fit_prior(prim_noabx, 0, 1))

# ---- Risk of bias (Newcastle-Ottawa) and pre-specified exclusion sensitivity ----
nos <- read_csv("nos_assessment.csv", show_col_types = FALSE)
low_rob <- nos$study_id[nos$rob_overall == "Low"]
n_low   <- length(low_rob)
# Protocol pre-specification: exclude studies at higher risk of bias (NOS < 7)
prim_abx_lr   <- prim_abx   |> filter(study_id %in% low_rob)
prim_noabx_lr <- prim_noabx |> filter(study_id %in% low_rob)
sens_prim <- summ(fit_prior(prim_abx_lr, 0, 1))
sens_sec  <- summ(fit_prior(prim_noabx_lr, 0, 1))

# ---- I-squared with posterior uncertainty (primary set) ----
# I2 = tau^2 / (tau^2 + s2_typ), with the Higgins-Thompson "typical" sampling variance
w_pri  <- 1 / prim_abx$sei^2
s2_typ <- (nrow(prim_abx) - 1) * sum(w_pri) / (sum(w_pri)^2 - sum(w_pri^2))
draws  <- fits_abx$primary$rposterior(20000)
i2     <- draws[, "tau"]^2 / (draws[, "tau"]^2 + s2_typ)
i2_txt <- sprintf("%.0f%% (95%% CrI %.0f\u2013%.0f%%)",
                  100 * median(i2), 100 * quantile(i2, 0.025), 100 * quantile(i2, 0.975))

# ---- Adjusted-only sensitivity: secondary (no-antibiotic) set ----
sec_adj_df   <- prim_noabx |> filter(adjusted)
sens_sec_adj <- summ(fit_prior(sec_adj_df, 0, 1))

# ---- Sensitivity excluding RR/HR/IRR-derived estimates converted to OR ----
# Per the extraction (Estimates sheet, Estimate_type of the curated estimate):
# Doernberg 2012 (aHR), Dial 2008 (aRR), Brown 2016 (IRR)
nonor <- c("Doernberg2012", "Dial2008", "Brown2016")
sens_prim_or <- summ(fit_prior(prim_abx   |> filter(!study_id %in% nonor), 0, 1))
sens_sec_or  <- summ(fit_prior(prim_noabx |> filter(!study_id %in% nonor), 0, 1))

# ---- Display helpers ----
fmt_ci <- function(x, lo, hi) sprintf("%.2f (%.2f\u2013%.2f)", x, lo, hi)
getp   <- function(p) as.list(res_abx[res_abx$prior == p, ])
P <- getp("primary")

# ---- Illustrative absolute risks for the GRADE summary-of-findings table ----
or2risk <- function(or, p0) 1000 * or * p0 / (1 - p0 + or * p0)
abs_prim <- sprintf("%.0f vs 20 per 1,000 (%.0f\u2013%.0f)",
                    or2risk(P$OR, 0.02), or2risk(P$OR_lo, 0.02), or2risk(P$OR_hi, 0.02))
abs_sec  <- sprintf("%.0f vs 5 per 1,000 (%.0f\u2013%.0f)",
                    or2risk(sec$OR, 0.005), or2risk(sec$OR_lo, 0.005), or2risk(sec$OR_hi, 0.005))

# ---- Subgroups within the primary set ----
prim_abx2 <- prim_abx |>
  mutate(design_broad = if_else(str_detect(study_design, "cohort"), "cohort", "case_control"))
subs <- list(
  doxycycline  = prim_abx2 |> filter(exposure_class == "doxycycline"),
  case_control = prim_abx2 |> filter(design_broad == "case_control"),
  cohort       = prim_abx2 |> filter(design_broad == "cohort")
)
res_sub <- imap_dfr(subs, ~ summ(fit_prior(.x, 0, 1)) |>
                      mutate(analysis = .y, k = nrow(.x), .before = 1))

# ---- Small-study effects (frequentist diagnostics) ----
rma_abx <- rma(yi = prim_abx$yi, vi = prim_abx$sei^2, method = "REML")
egg <- regtest(rma_abx)                       # Egger's regression test

# ---- Leave-one-out on the Bayesian primary (prior N(0,1)) ----
loo <- map_dfr(seq_len(nrow(prim_abx)), function(i) {
  d <- prim_abx[-i, ]
  s <- summ(bayesmeta(y = d$yi, sigma = d$sei, labels = d$reference,
                      mu.prior.mean = 0, mu.prior.sd = 1, tau.prior = hn))
  tibble(dropped = prim_abx$reference[i],
         OR = s$OR, OR_lo = s$OR_lo, OR_hi = s$OR_hi, P1 = s$P_OR_lt_1)
})

# ---- Comparison with Tariq (2018): recapitulation & decomposition ----
# Tariq's exact per-study estimates (Figure 2A of Tariq et al. 2018)
tariq <- tibble(
  study = c("Baxter 2008","Delaney 2007","Dial 2008","Doernberg 2012","Kuntz 2011","Tartof 2015"),
  yi  = c(-0.8916, -0.1054, 0.0953, -0.3147, -0.0619, -0.6931),
  sei = c( 0.2324,  0.2999, 1.2234,  0.1353,  0.3990,  0.1139))
dl <- rma(yi = tariq$yi, sei = tariq$sei, method = "DL")          # reproduce Tariq's DL model
bm_tariq  <- summ(bayesmeta(tariq$yi, tariq$sei,
                            mu.prior.mean = 0, mu.prior.sd = 1, tau.prior = hn))
tariq_ids <- c("Baxter2008","Delaney2007","Dial2008","Doernberg2012","Kuntz2011","Tartof2015")
new8 <- prim_abx |> filter(!study_id %in% tariq_ids)             # new antibiotic-comparator studies
bm_bridge <- summ(bayesmeta(c(tariq$yi, new8$yi), c(tariq$sei, new8$sei),
                            mu.prior.mean = 0, mu.prior.sd = 1, tau.prior = hn))
comp <- bind_rows(
  tibble(analysis = "Tariq 2018, original (DL frequentist)", k = 6,
         OR = exp(coef(dl)), OR_lo = exp(dl$ci.lb), OR_hi = exp(dl$ci.ub)),
  bm_tariq  |> transmute(analysis = "Tariq studies, our Bayesian model", k = 6, OR, OR_lo, OR_hi),
  bm_bridge |> transmute(analysis = "Tariq studies + 8 new studies", k = 14, OR, OR_lo, OR_hi),
  getp("primary") |> as_tibble() |>
    transmute(analysis = "This review, primary (Bayesian)", k = 11, OR, OR_lo, OR_hi))

v_tariq_dl  <- fmt_ci(exp(coef(dl)), exp(dl$ci.lb), exp(dl$ci.ub))
v_tariq_bay <- fmt_ci(bm_tariq$OR, bm_tariq$OR_lo, bm_tariq$OR_hi)
v_tariq_p   <- sprintf("%.2f", bm_tariq$P_OR_lt_1)
v_bridge    <- fmt_ci(bm_bridge$OR, bm_bridge$OR_lo, bm_bridge$OR_hi)
v_bridge_p  <- sprintf("%.2f", bm_bridge$P_OR_lt_1)

# ---- Care-context meta-regression (exact Bayesian, bmr) ----
# PRIMARY: antibiotic-comparator set (k = 11)
Xc <- model.matrix(~ 0 + care_context, data = prim_abx); colnames(Xc) <- c("CA","HA","mixed")
mr <- bmr(y = prim_abx$yi, sigma = prim_abx$sei, labels = prim_abx$reference, X = Xc,
          tau.prior = hn, beta.prior.mean = c(0,0,0), beta.prior.sd = c(1,1,1))
ctx <- tibble(care_context = c("CA","HA","mixed"),
  k     = as.integer(colSums(Xc)),
  OR    = exp(mr$summary["median",    c("CA","HA","mixed")]),
  OR_lo = exp(mr$summary["95% lower", c("CA","HA","mixed")]),
  OR_hi = exp(mr$summary["95% upper", c("CA","HA","mixed")]),
  P_OR_lt_1 = c(mr$pposterior(beta = 0, which = "CA"),
                mr$pposterior(beta = 0, which = "HA"),
                mr$pposterior(beta = 0, which = "mixed")))
mr_tau <- sprintf("%.2f", mr$summary["median","tau"])
# SENSITIVITY: full primary set (k = 20)
Xc20 <- model.matrix(~ 0 + care_context, data = prim); colnames(Xc20) <- c("CA","HA","mixed")
mr20 <- bmr(y = prim$yi, sigma = prim$sei, labels = prim$reference, X = Xc20,
            tau.prior = hn, beta.prior.mean = c(0,0,0), beta.prior.sd = c(1,1,1))
ctx20 <- tibble(care_context = c("CA","HA","mixed"),
  k     = as.integer(colSums(Xc20)),
  OR    = exp(mr20$summary["median",    c("CA","HA","mixed")]),
  OR_lo = exp(mr20$summary["95% lower", c("CA","HA","mixed")]),
  OR_hi = exp(mr20$summary["95% upper", c("CA","HA","mixed")]),
  P_OR_lt_1 = c(mr20$pposterior(beta = 0, which = "CA"),
                mr20$pposterior(beta = 0, which = "HA"),
                mr20$pposterior(beta = 0, which = "mixed")))
ha20   <- fmt_ci(ctx20$OR[2], ctx20$OR_lo[2], ctx20$OR_hi[2])
ha20_p <- sprintf("%.2f", ctx20$P_OR_lt_1[2])

# ---- Design & adjustment meta-regressions (pre-specified §19; exact Bayesian, bmr) ----
# Design: primary antibiotic-comparator set (k = 11), collapsed case-control vs cohort.
prim_abx <- prim_abx |>
  mutate(design2 = if_else(str_detect(study_design, "case_control"), "case_control", "cohort"))
Xd  <- model.matrix(~ 0 + design2, data = prim_abx); colnames(Xd) <- c("case_control","cohort")
mrd <- bmr(y = prim_abx$yi, sigma = prim_abx$sei, labels = prim_abx$reference, X = Xd,
           tau.prior = hn, beta.prior.mean = c(0, 0), beta.prior.sd = c(1, 1))
dsn <- tibble(group = c("Case-control","Cohort"),
  k         = as.integer(colSums(Xd)),
  OR        = exp(mrd$summary["median",    c("case_control","cohort")]),
  OR_lo     = exp(mrd$summary["95% lower", c("case_control","cohort")]),
  OR_hi     = exp(mrd$summary["95% upper", c("case_control","cohort")]),
  P_OR_lt_1 = c(mrd$pposterior(beta = 0, which = "case_control"),
                mrd$pposterior(beta = 0, which = "cohort")))
mrd_tau <- sprintf("%.2f", mrd$summary["median","tau"])

# Adjustment: all 11 primary-abx estimates are adjusted, so this covariate is assessed on
# the full primary set (k = 20; 14 adjusted, 6 unadjusted).
prim <- prim |> mutate(adj2 = if_else(as.logical(adjusted), "adjusted", "unadjusted"))
Xa  <- model.matrix(~ 0 + adj2, data = prim); colnames(Xa) <- c("adjusted","unadjusted")
mra <- bmr(y = prim$yi, sigma = prim$sei, labels = prim$reference, X = Xa,
           tau.prior = hn, beta.prior.mean = c(0, 0), beta.prior.sd = c(1, 1))
adjt <- tibble(group = c("Adjusted","Unadjusted"),
  k         = as.integer(colSums(Xa)),
  OR        = exp(mra$summary["median",    c("adjusted","unadjusted")]),
  OR_lo     = exp(mra$summary["95% lower", c("adjusted","unadjusted")]),
  OR_hi     = exp(mra$summary["95% upper", c("adjusted","unadjusted")]),
  P_OR_lt_1 = c(mra$pposterior(beta = 0, which = "adjusted"),
                mra$pposterior(beta = 0, which = "unadjusted")))
mra_tau <- sprintf("%.2f", mra$summary["median","tau"])

# ---- Prior-data conflict: literature prior vs data-only estimate (Box / Marshall-Spiegelhalter) ----
fit_vague_pdc <- bayesmeta(prim_abx$yi, prim_abx$sei, mu.prior.mean = 0, mu.prior.sd = 10, tau.prior = hn)
pdc_md <- fit_vague_pdc$summary["mean","mu"]; pdc_sd <- fit_vague_pdc$summary["sd","mu"]
pdc_m0 <- log(0.62); pdc_s0 <- 0.10
pdc_z  <- (pdc_m0 - pdc_md) / sqrt(pdc_s0^2 + pdc_sd^2)
pdc_p  <- 2 * pnorm(-abs(pdc_z))

# ---- JAMA-style theme and blue-gray palette ----
jama_navy <- "#2C4A63"; jama_blue <- "#3B6E9C"; jama_gray <- "#5A6B7B"; jama_band <- "#CBD8E3"
theme_jama <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(text = element_text(colour = "#1A1A1A"),
          axis.line = element_line(colour = "#4D4D4D", linewidth = 0.4),
          axis.ticks = element_line(colour = "#4D4D4D", linewidth = 0.4),
          panel.grid.major.x = element_line(colour = "#EAEDF0", linewidth = 0.3),
          plot.title = element_text(face = "bold", size = rel(1.02)),
          plot.subtitle = element_text(colour = "#4D4D4D", size = rel(0.9)),
          legend.position = "top", legend.title = element_blank())
}
```

```{r toggle-assets}
#| output: asis
cat('
<style>
  body.lang-en .it { display: none; }
  body.lang-it .en { display: none; }
  .lang-toggle { display:flex; gap:0; border:1px solid #2c7fb8; border-radius:999px;
    overflow:hidden; width:max-content; margin:0 0 1.5rem; }
  .lang-toggle button { background:#fff; color:#2c7fb8; border:0; padding:.45rem 1rem;
    cursor:pointer; font-size:.9rem; font-weight:600; }
  .lang-toggle button.active { background:#2c7fb8; color:#fff; }
  .card { background:#f2f7fb; border:1px solid #d4e3ef; border-radius:12px;
    padding:1rem 1.25rem; margin:1rem 0; }
  .big { font-size:1.6rem; font-weight:700; color:#2c7fb8; }
  .cap { color:#5a6b7a; font-size:.88rem; margin:.3rem 0 1.4rem; }
</style>
<div class="lang-toggle" role="group" aria-label="Language">
  <button id="btn-en" class="active" onclick="setLang(\'en\')">English</button>
  <button id="btn-it" onclick="setLang(\'it\')">Italiano</button>
</div>
<script>
  function setLang(l){
    document.body.classList.remove("lang-en","lang-it");
    document.body.classList.add("lang-"+l);
    document.documentElement.lang = l;
    document.getElementById("btn-en").classList.toggle("active", l==="en");
    document.getElementById("btn-it").classList.toggle("active", l==="it");
  }
  document.addEventListener("DOMContentLoaded", function(){ setLang("en"); });
</script>
')
```

## [Background]{.en}[Contesto]{.it}

::: {.en}
*Clostridioides difficile* infection (CDI) is among the most common healthcare-associated
infections and a leading cause of antibiotic-associated diarrhoea and colitis. Systemic
antibiotic exposure is the principal modifiable risk factor: by disrupting the protective
gut microbiota, antibiotics allow *C. difficile* to proliferate and release toxin. That risk
is not uniform across drug classes — agents such as clindamycin, later-generation
cephalosporins and fluoroquinolones are consistently linked to higher CDI risk — so the
choice of antibiotic is itself a lever for prevention and antimicrobial stewardship.

Tetracyclines, and doxycycline in particular, have been proposed as comparatively
lower-risk agents. The mechanistic rationale is a relatively narrow impact on the anaerobic
colonic flora that resists colonisation, together with possible direct suppression of
*C. difficile* growth and toxin production. A 2018 systematic review and meta-analysis
pooled six observational studies and reported roughly a 38% lower risk of CDI with
tetracycline exposure (OR ≈ 0.62) [@Tariq2018]. That estimate, however, combined studies
with differing comparators (other antibiotics and, in some studies, no antibiotic), rested
on a modest evidence base, and predates a substantial body of subsequent observational data,
including several very large administrative-cohort studies.

This report updates and re-examines the question under a pre-registered protocol. It
analyses the comparison the protocol specifies — tetracyclines versus *other antibiotics* —
separately from the distinct question of tetracyclines versus *no antibiotic*, and estimates
the association within a Bayesian random-effects framework that makes the residual
heterogeneity and uncertainty explicit.

**Aim.** To determine whether, and with what certainty, tetracycline exposure is associated
with a lower risk of incident CDI than other antibiotics — the question posed formally below.
:::
::: {.it}
L'infezione da *Clostridioides difficile* (ICD) è tra le infezioni correlate
all'assistenza più comuni e una delle principali cause di diarrea e colite associate agli
antibiotici. L'esposizione ad antibiotici sistemici è il principale fattore di rischio
modificabile: alterando il microbiota intestinale protettivo, gli antibiotici consentono a
*C. difficile* di proliferare e rilasciare tossina. Questo rischio non è uniforme tra le
classi di farmaci — agenti come clindamicina, cefalosporine di ultima generazione e
fluorochinoloni sono costantemente associati a un rischio di ICD più elevato — per cui la
scelta dell'antibiotico è essa stessa una leva per la prevenzione e la stewardship
antimicrobica.

Le tetracicline, e in particolare la doxiciclina, sono state proposte come agenti a rischio
comparativamente inferiore. La motivazione meccanicistica è un impatto relativamente
ristretto sulla flora colica anaerobia che resiste alla colonizzazione, insieme a una
possibile soppressione diretta della crescita e della produzione di tossina di
*C. difficile*. Una revisione sistematica con meta-analisi del 2018 ha aggregato sei studi
osservazionali riportando un rischio di ICD inferiore di circa il 38% con l'esposizione alle
tetracicline (OR ≈ 0,62) [@Tariq2018]. Quella stima, tuttavia, combinava studi con
comparatori diversi (altri antibiotici e, in alcuni studi, nessun antibiotico), poggiava su
una base di evidenze modesta e precede un consistente corpus di dati osservazionali
successivi, tra cui diversi studi di coorte amministrativi molto ampi.

Questo report aggiorna e riesamina la questione con un protocollo pre-registrato. Analizza
il confronto specificato dal protocollo — tetracicline versus *altri antibiotici* — in modo
separato dalla domanda distinta delle tetracicline versus *nessun antibiotico*, e stima
l'associazione con un modello bayesiano a effetti casuali che rende esplicite l'eterogeneità
e l'incertezza residue.

**Obiettivo.** Stabilire se, e con quale grado di certezza, l'esposizione alle tetracicline
sia associata a un rischio di ICD incidente inferiore rispetto ad altri antibiotici — la
domanda posta formalmente qui sotto.
:::

## [Research question]{.en}[Domanda di ricerca]{.it}

::: {.en}
Among patients exposed to systemic antibiotics, is exposure to a tetracycline-class
antibiotic associated with a different risk of incident *C. difficile* infection (CDI)
**compared with other antibiotics**? This report re-estimates the association within a
Bayesian random-effects framework and quantifies the posterior probability of a
protective effect. Studies whose only comparator was "no antibiotic" are analysed
separately as a secondary comparison, per the PROSPERO protocol's PECO.
:::

::: {.it}
Nei pazienti esposti ad antibiotici sistemici, l'esposizione a un antibiotico della
classe delle tetracicline è associata a un rischio diverso di infezione incidente da
*C. difficile* (ICD) **rispetto ad altri antibiotici**? Questo report ristima
l'associazione con un modello bayesiano a effetti casuali e quantifica la probabilità a
posteriori di un effetto protettivo. Gli studi il cui unico comparatore era "nessun
antibiotico" sono analizzati separatamente come confronto secondario, secondo il PECO
del protocollo PROSPERO.
:::

## [Study selection (PRISMA 2020)]{.en}[Selezione degli studi (PRISMA 2020)]{.it}

![](report_figs/prisma2020_flow){fig-align="center" width="100%"}

::: {.cap}
::: {.en}
**PRISMA 2020 flow.** Canonical PRISMA 2020 diagram (generated with the `PRISMA2020` R
package). Identification and de-duplication counts are validated (2,926 database + 1
register records + 95 from citation searching of Tariq 2018; 1,452 duplicates removed).
Screening-stage counts are *provisional*, reconstructed from the Elicit AI second-screen
pending human dual screening per the protocol (`PRISMA_flow_data.xlsx`): databases/registers
arm 1,475 screened → 40 assessed → 11 full-text exclusions; the 95 other-methods (citation)
reports yielded no additional eligible study. Of the 35 included studies (29 new + 6 carried
forward from Tariq 2018), 20 contribute a curated effect size to the Bayesian meta-analysis:
**11 in the primary comparison** (tetracyclines vs other antibiotics) and **9 in the
secondary comparison** (tetracyclines vs no antibiotic).
:::
::: {.it}
**Flusso PRISMA 2020.** Diagramma PRISMA 2020 canonico (generato con il pacchetto R
`PRISMA2020`). I conteggi di identificazione e deduplicazione sono validati (2.926 record da
database + 1 registro + 95 da citation searching di Tariq 2018; 1.452 duplicati rimossi). I
conteggi della fase di screening sono *provvisori*, ricostruiti dal secondo screening AI di
Elicit in attesa del doppio screening manuale secondo il protocollo (`PRISMA_flow_data.xlsx`):
braccio database/registri 1.475 sottoposti a screening → 40 valutati → 11 esclusioni full-text;
i 95 report da altri metodi (citation searching) non hanno prodotto ulteriori studi eleggibili.
Dei 35 studi inclusi (29 nuovi + 6 trasferiti da Tariq 2018), 20 contribuiscono con una stima
curata alla meta-analisi bayesiana: **11 nel confronto primario** (tetracicline vs altri
antibiotici) e **9 nel confronto secondario** (tetracicline vs nessun antibiotico).
:::
:::

## [Inter-rater agreement (screening)]{.en}[Concordanza tra revisori (screening)]{.it}

::: {.en}
The protocol (§16) requires two reviewers to screen titles/abstracts independently, with
agreement reported as Cohen's κ. Human dual screening is **not yet complete**, so a
**synthetic, illustrative** decision log (`screening_decisions.csv`, every row tagged
`SYNTHETIC_ILLUSTRATIVE`) is included here purely to demonstrate the calculation. The κ below
is therefore illustrative only and must be replaced by the real two-reviewer log (identical
schema: columns `reviewer1`/`reviewer2` holding include/exclude per record); the chunk then
reports the true agreement automatically.
:::
::: {.it}
Il protocollo (§16) richiede che due revisori effettuino lo screening di titoli/abstract in
modo indipendente, riportando la concordanza come κ di Cohen. Il doppio screening manuale
**non è ancora completo**, quindi qui è incluso un registro decisionale **sintetico e
illustrativo** (`screening_decisions.csv`, con ogni riga contrassegnata da
`SYNTHETIC_ILLUSTRATIVE`) al solo scopo di dimostrare il calcolo. Il κ sottostante è quindi
solo illustrativo e deve essere sostituito dal registro reale dei due revisori (schema
identico: colonne `reviewer1`/`reviewer2` con include/exclude per record); il blocco riporterà
allora la concordanza reale automaticamente.
:::

```{r kappa, results='asis'}
#| cache: false
kappa_file <- "screening_decisions.csv"
if (file.exists(kappa_file)) {
  kd  <- readr::read_csv(kappa_file, show_col_types = FALSE)
  tab <- table(kd$reviewer1, kd$reviewer2)
  po  <- sum(diag(tab)) / sum(tab)                       # observed agreement
  pe  <- sum(rowSums(tab) * colSums(tab)) / sum(tab)^2   # chance agreement
  kap <- (po - pe) / (1 - pe)                            # Cohen's kappa
  synth <- "source" %in% names(kd) && any(kd$source == "SYNTHETIC_ILLUSTRATIVE")
  cat(sprintf(
    "**Cohen's \u03ba = %.2f** (observed agreement %.1f%%; n = %d records)%s.",
    kap, 100 * po, sum(tab),
    if (synth) " \u2014 **illustrative synthetic data; replace with the real two-reviewer log**" else ""))
} else {
  cat("*Pending human dual screening \u2014 no `screening_decisions.csv` in the project yet.*")
}
```

## [Table 1. Characteristics of included studies]{.en}[Tabella 1. Caratteristiche degli studi inclusi]{.it}

```{r tbl1, results='asis'}
prim |>
  arrange(exp(yi)) |>
  transmute(
    Study = sprintf("[@%s]", study_id),
    Design = str_replace_all(study_design, "_", " "),
    Setting = care_context,
    N = ifelse(is.na(ni), "NR", formatC(ni, format = "d", big.mark = ",")),
    Agent = str_replace_all(exposure_class, "_", " "),
    Comparator = str_replace_all(comparator_type, "_", " "),
    `Adj.` = ifelse(adjusted, "Yes", "No"),
    `OR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", exp(yi), exp(yi - 1.96*sei), exp(yi + 1.96*sei))) |>
  knitr::kable(format = "pipe", align = "lllrllcr") |>
  cat(sep = "\n")
```

::: {.cap}
::: {.en}
**Table 1.** Characteristics of the 20 studies contributing to the Bayesian synthesis
(11 vs other antibiotics; 9 vs no antibiotic; ordered by odds ratio). N = analytic sample
size (NR = not reported); Adj. = adjusted estimate. The OR is the single curated effect size
carried into the meta-analysis; each study is identified by its bibliographic citation.
:::
::: {.it}
**Tabella 1.** Caratteristiche dei 20 studi nella sintesi bayesiana (11 vs altri antibiotici;
9 vs nessun antibiotico; ordinati per odds ratio). N = dimensione campionaria analitica
(NR = non riportata); Adj. = stima aggiustata. L'OR è la singola stima d'effetto curata
usata nella meta-analisi; ogni studio è identificato dalla relativa citazione bibliografica.
:::
:::

## [Risk of bias (Newcastle-Ottawa Scale)]{.en}[Rischio di bias (scala Newcastle-Ottawa)]{.it}

::: {.en}
All 20 studies contributing to the synthesis are observational, so risk of bias was
assessed with the Newcastle-Ottawa Scale (NOS) [@WellsNOS], using the design-appropriate
version (cohort or case-control; maximum 9 stars across the Selection, Comparability and
Outcome/Exposure domains). Uniform decision rules were applied: the case-definition item
was credited only for laboratory-confirmed or independently validated definitions
(unvalidated ICD/claims algorithms were not); comparability stars required matching or
adjustment for age/sex (first star) and for additional key confounders (second star), so
studies whose curated estimate is unadjusted beyond matching received one comparability
star; for cohorts, completeness of follow-up was credited only for closed healthcare
systems (HMO, claims or population registers). Overall ratings: 7-9 stars = **low** risk
of bias, 5-6 = **some concerns**, 4 or fewer = **high**. Per-item scores and
study-level justifications are recorded in `nos_assessment.csv`; the study-level judgements
are summarised as a traffic-light plot in Figure 9.
:::
::: {.it}
Tutti i 20 studi che contribuiscono alla sintesi sono osservazionali; il rischio di bias è
stato valutato con la scala Newcastle-Ottawa (NOS) [@WellsNOS], nella versione appropriata
al disegno (coorte o caso-controllo; massimo 9 stelle nei domini Selezione, Comparabilità
ed Esito/Esposizione). Sono state applicate regole decisionali uniformi: la definizione di
caso è stata accreditata solo se confermata in laboratorio o basata su codici validati
(gli algoritmi ICD/claims non validati non ricevono la stella); le stelle di comparabilità
richiedono appaiamento o aggiustamento per età/sesso (prima stella) e per ulteriori
confondenti chiave (seconda stella), per cui gli studi la cui stima curata non è
aggiustata oltre l'appaiamento ricevono una sola stella; per le coorti, la completezza del
follow-up è stata accreditata solo per sistemi sanitari chiusi (HMO, claims o registri di
popolazione). Valutazioni complessive: 7-9 stelle = rischio di bias **basso**, 5-6 =
**alcune perplessità**, 4 o meno = **alto**. I punteggi per singolo item e le
giustificazioni per studio sono registrati in `nos_assessment.csv`; i giudizi per studio sono
riassunti come traffic-light plot nella Figura 9.
:::

```{r rob-table}
#| results: asis
nos |>
  arrange(desc(total), study) |>
  transmute(
    Study = sprintf("[@%s]", study_id),
    Design = design,
    Set = analysis_set,
    `Selection (0-4)` = sel1 + sel2 + sel3 + sel4,
    `Comparability (0-2)` = comp1 + comp2,
    `Outcome/Exposure (0-3)` = oe1 + oe2 + oe3,
    `Total (0-9)` = total,
    `Overall RoB` = rob_overall) |>
  knitr::kable(format = "pipe", align = "lllrrrrl") |>
  cat(sep = "\n")
```

::: {.cap}
::: {.en}
**Table 2.** Newcastle-Ottawa assessment of the 20 studies in the synthesis (domain
subtotals; maxima 4/2/3 stars, 9 overall). Selection covers case/cohort definition,
representativeness, comparison-group selection and absence of the outcome at baseline
(definition of controls for case-control studies); Outcome/Exposure covers ascertainment
of outcome (cohorts) or exposure (case-control studies), comparability of ascertainment
across groups, and follow-up/non-response.
:::
::: {.it}
**Tabella 2.** Valutazione Newcastle-Ottawa dei 20 studi della sintesi (subtotali per
dominio; massimi 4/2/3 stelle, 9 complessive). La selezione copre definizione di
caso/coorte, rappresentatività, selezione del gruppo di confronto e assenza dell'esito al
baseline (definizione dei controlli per i caso-controllo); Esito/Esposizione copre
l'accertamento dell'esito (coorti) o dell'esposizione (caso-controllo), la comparabilità
dell'accertamento tra i gruppi e il follow-up/non-risposta.
:::
:::

```{r rob-plot}
#| fig-width: 6.5
#| fig-height: 7
rob_dat <- nos |>
  mutate(
    Selection = case_when(sel1 + sel2 + sel3 + sel4 >= 3 ~ "Low",
                          sel1 + sel2 + sel3 + sel4 == 2 ~ "Some concerns", TRUE ~ "High"),
    Comparability = case_when(comp1 + comp2 == 2 ~ "Low",
                              comp1 + comp2 == 1 ~ "Some concerns", TRUE ~ "High"),
    `Outcome / Exposure` = case_when(oe1 + oe2 + oe3 == 3 ~ "Low",
                                     oe1 + oe2 + oe3 == 2 ~ "Some concerns", TRUE ~ "High"),
    Overall = rob_overall) |>
  pivot_longer(c(Selection, Comparability, `Outcome / Exposure`, Overall),
               names_to = "domain", values_to = "rating") |>
  mutate(domain = factor(domain, levels = c("Selection", "Comparability",
                                            "Outcome / Exposure", "Overall")),
         rating = factor(rating, levels = c("Low", "Some concerns", "High")),
         study  = reorder(study, total))

# ggplot2 (>= 4.0) draws no legend key for a factor level absent from the
# data; a "High" tile hidden beneath a real one keeps the red key visible
rob_ghost <- rob_dat[1, ] |>
  mutate(rating = factor("High", levels = c("Low", "Some concerns", "High")))

ggplot(rob_dat, aes(x = domain, y = study, fill = rating)) +
  geom_tile(data = rob_ghost, colour = "white", linewidth = 0.8) +
  geom_tile(colour = "white", linewidth = 0.8) +
  scale_fill_manual(values = c("Low" = "#4C9A62", "Some concerns" = "#E9B44C",
                               "High" = "#C0504D"), drop = FALSE) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), legend.position = "bottom")
```

::: {.cap}
::: {.en}
**Traffic-light plot.** Domain-level risk-of-bias ratings. Domain rules: Selection low if
at least 3 of 4 stars; Comparability low with 2 stars, some concerns with 1;
Outcome/Exposure low with 3 stars, some concerns with 2. Overall: 7-9 low, 5-6 some
concerns, 4 or fewer high.
:::
::: {.it}
**Traffic-light plot.** Valutazioni di rischio di bias per dominio. Regole: Selezione
bassa con almeno 3 stelle su 4; Comparabilità bassa con 2 stelle, alcune perplessità con
1; Esito/Esposizione bassa con 3 stelle, alcune perplessità con 2. Complessivo: 7-9
basso, 5-6 alcune perplessità, 4 o meno alto.
:::
:::

::: {.en}
`r n_low` of 20 studies reach 7 or more stars (low risk of bias). Recurring limitations
are case definitions based on unvalidated diagnostic codes (selection), hospital-based
controls in smaller case-control studies (selection), reliance on unadjusted per-drug
estimates (comparability), and uncertain completeness of post-discharge outcome capture
in hospital-discharge cohorts (outcome). For the six studies carried forward from Tariq
et al. (2018), our independent scores (median 8, range 7-9) are consistent with theirs
(median 7, range 7-8) [@Tariq2018].

The pre-specified sensitivity analysis excluding studies at higher risk of bias (NOS
below 7) removes only Predrag 2016: the primary estimate is unchanged because all 11
antibiotic-comparator studies are low risk (OR
`r fmt_ci(sens_prim$OR, sens_prim$OR_lo, sens_prim$OR_hi)`), and the secondary,
no-antibiotic comparison becomes OR
`r fmt_ci(sens_sec$OR, sens_sec$OR_lo, sens_sec$OR_hi)` (k = 8). Conclusions are
therefore robust to study quality as assessed here.
:::
::: {.it}
`r n_low` studi su 20 raggiungono almeno 7 stelle (basso rischio di bias). Le limitazioni
ricorrenti sono definizioni di caso basate su codici diagnostici non validati
(selezione), controlli ospedalieri negli studi caso-controllo più piccoli (selezione),
affidamento a stime per singolo farmaco non aggiustate (comparabilità) e incerta
completezza della cattura dell'esito dopo la dimissione nelle coorti ospedaliere (esito).
Per i sei studi trasferiti da Tariq et al. (2018), i nostri punteggi indipendenti
(mediana 8, intervallo 7-9) sono coerenti con i loro (mediana 7, intervallo 7-8)
[@Tariq2018].

L'analisi di sensibilità pre-specificata che esclude gli studi a maggior rischio di bias
(NOS inferiore a 7) rimuove solo Predrag 2016: la stima primaria è invariata perché tutti
gli 11 studi con comparatore antibiotico sono a basso rischio (OR
`r fmt_ci(sens_prim$OR, sens_prim$OR_lo, sens_prim$OR_hi)`), e il confronto secondario
(vs nessun antibiotico) diventa OR
`r fmt_ci(sens_sec$OR, sens_sec$OR_lo, sens_sec$OR_hi)` (k = 8). Le conclusioni sono
quindi robuste rispetto alla qualità degli studi così come valutata qui.
:::


::: {.callout-note}
::: {.en}
**Protocol deviation — no randomised trials included.** The protocol pre-specified RCTs as an
eligible design, with a study-design stratification (observational vs RCT) and a Cochrane
**RoB 2** appraisal for any trials. No RCT met the inclusion criteria: all 20 studies in the
synthesis are observational. The observational-vs-RCT stratum and the RoB 2 appraisal are
therefore not applicable and were not performed, and the design meta-regression is restricted
to observational designs (case-control vs cohort).
:::
::: {.it}
**Deviazione dal protocollo — nessuno studio randomizzato incluso.** Il protocollo prevedeva
gli RCT tra i disegni eleggibili, con una stratificazione per disegno (osservazionale vs RCT) e
la valutazione Cochrane **RoB 2** per eventuali trial. Nessun RCT ha soddisfatto i criteri di
inclusione: tutti i 20 studi della sintesi sono osservazionali. Lo strato osservazionale-vs-RCT
e la valutazione RoB 2 non sono quindi applicabili e non sono stati eseguiti, e la
meta-regressione per disegno è limitata ai disegni osservazionali (caso-controllo vs coorte).
:::
:::

## [Methods]{.en}[Metodi]{.it}

::: {.en}
Hierarchical random-effects model on the log-odds scale:
$$y_i \sim \text{Normal}(\theta_i, s_i^2), \qquad \theta_i \sim \text{Normal}(\mu, \tau^2).$$
Primary prior $\mu \sim \text{Normal}(0, 1^2)$; heterogeneity prior
$\tau \sim \text{Half-Normal}(0, 0.5)$. Posteriors were computed exactly by numerical
integration (`bayesmeta`) and independently confirmed by Hamiltonian Monte Carlo
(`brms`/Stan, 4 chains). We report the pooled odds ratio (OR = exp $\mu$) with a 95%
credible interval (CrI), the between-study SD $\tau$, the predictive interval for a new
study, and the posterior probabilities P(OR<1) and P(OR<0.80). A pre-specified
prior-sensitivity analysis used vague, sceptical and literature-informed priors.
:::

::: {.it}
Modello gerarchico a effetti casuali su scala log-odds:
$$y_i \sim \text{Normale}(\theta_i, s_i^2), \qquad \theta_i \sim \text{Normale}(\mu, \tau^2).$$
Prior primario $\mu \sim \text{Normale}(0, 1^2)$; prior sull'eterogeneità
$\tau \sim \text{Half-Normale}(0, 0.5)$. Le distribuzioni a posteriori sono state
calcolate in modo esatto per integrazione numerica (`bayesmeta`) e confermate in modo
indipendente con Hamiltonian Monte Carlo (`brms`/Stan, 4 catene). Riportiamo l'odds ratio
aggregato (OR = exp $\mu$) con intervallo di credibilità (ICr) al 95%, la deviazione
standard tra studi $\tau$, l'intervallo predittivo per un nuovo studio e le probabilità a
posteriori P(OR<1) e P(OR<0,80). Un'analisi di sensibilità sui prior pre-specificata ha
usato prior vago, scettico e informato dalla letteratura.
:::

## [Primary result (11 studies)]{.en}[Risultato primario (11 studi)]{.it}

::: {.card}
::: {.en}
Pooled odds ratio (tetracyclines vs other antibiotics)
<span class="big">`r fmt_ci(P$OR, P$OR_lo, P$OR_hi)`</span>

The 95% credible interval crosses 1. The posterior probability of any protective effect
is **P(OR<1) = `r sprintf("%.2f", P$P_OR_lt_1)`**, and the probability of a clinically
meaningful effect is **P(OR<0.80) = `r sprintf("%.2f", P$P_OR_lt_080)`**. Between-study
heterogeneity is substantial ($\tau$ = `r fmt_ci(P$tau, P$tau_lo, P$tau_hi)`;
I² = `r i2_txt`): the
predictive interval for a future study spans OR
`r sprintf("%.2f\u2013%.2f", P$pred_lo, P$pred_hi)`.
:::
::: {.it}
Odds ratio aggregato (tetracicline vs altri antibiotici)
<span class="big">`r fmt_ci(P$OR, P$OR_lo, P$OR_hi)`</span>

L'intervallo di credibilità al 95% attraversa 1. La probabilità a posteriori di un
qualsiasi effetto protettivo è **P(OR<1) = `r sprintf("%.2f", P$P_OR_lt_1)`**, e la
probabilità di un effetto clinicamente rilevante è
**P(OR<0,80) = `r sprintf("%.2f", P$P_OR_lt_080)`**. L'eterogeneità tra studi è
sostanziale ($\tau$ = `r fmt_ci(P$tau, P$tau_lo, P$tau_hi)`; I² = `r i2_txt`):
l'intervallo predittivo per
uno studio futuro va da OR `r sprintf("%.2f\u2013%.2f", P$pred_lo, P$pred_hi)`.
:::
:::

```{r fig-forest, fig.width=8, fig.height=5.2, cache=TRUE}
pooled <- getp("primary")
fit_prim <- bayesmeta(prim_abx$yi, prim_abx$sei, labels = prim_abx$reference,
                      mu.prior.mean = 0, mu.prior.sd = 1, tau.prior = hn)
th  <- fit_prim$theta                                    # per-study posterior (shrinkage)
ord <- prim_abx |> arrange(exp(yi)) |> pull(reference)
fdf <- bind_rows(
  prim_abx |> transmute(study = reference, type = "Observed (study estimate)",
                        OR = exp(yi), lo = exp(yi - 1.96*sei), hi = exp(yi + 1.96*sei)),
  tibble(study = colnames(th), type = "Bayesian posterior (shrinkage)",
         OR = exp(th["median",]), lo = exp(th["95% lower",]), hi = exp(th["95% upper",]))
) |> mutate(study = factor(study, levels = ord),
            type = factor(type, levels = c("Observed (study estimate)",
                                           "Bayesian posterior (shrinkage)")))

priors_v <- tibble(
  label = factor(c("Sceptical prior", "Literature-informed prior"),
                 levels = c("Sceptical prior", "Literature-informed prior")),
  OR = c(res_abx$OR[res_abx$prior == "sceptical"], res_abx$OR[res_abx$prior == "literature"]))
sens_lo <- min(priors_v$OR, pooled$OR); sens_hi <- max(priors_v$OR, pooled$OR)

ggplot(fdf, aes(OR, study, colour = type)) +
  annotate("rect", xmin = sens_lo, xmax = sens_hi, ymin = -Inf, ymax = Inf,
           fill = jama_navy, alpha = 0.06) +
  annotate("rect", xmin = pooled$OR_lo, xmax = pooled$OR_hi, ymin = -Inf, ymax = Inf,
           fill = jama_band, alpha = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#4D4D4D", linewidth = 0.4) +
  geom_vline(xintercept = pooled$OR, colour = jama_navy, linewidth = 0.7) +
  geom_vline(data = priors_v, aes(xintercept = OR, linetype = label),
             colour = jama_navy, linewidth = 0.5) +
  geom_pointrange(aes(xmin = lo, xmax = hi), position = position_dodge(width = 0.6), size = 0.4) +
  scale_colour_manual(values = c("Observed (study estimate)" = jama_gray,
                                 "Bayesian posterior (shrinkage)" = jama_blue)) +
  scale_linetype_manual(values = c("Sceptical prior" = "dotted",
                                   "Literature-informed prior" = "longdash")) +
  scale_x_log10(breaks = c(0.1, 0.25, 0.5, 1, 2, 4, 10)) +
  labs(x = "Odds ratio (log scale)", y = NULL, linetype = "Pooled OR under prior") +
  guides(colour = guide_legend(order = 1), linetype = guide_legend(order = 2)) +
  theme_jama() + theme(legend.box = "vertical", legend.spacing.y = unit(1, "pt"))
```

::: {.cap}
::: {.en}
**Figure 1.** Study-level odds ratios (primary set, 11 studies with antibiotic comparators).
Gray = observed estimates; blue = Bayesian posterior (shrinkage) estimates, pulled toward
the pooled mean. The navy line and blue band show the pooled OR (primary prior) and 95%
credible interval. The dotted and long-dashed vertical lines mark the pooled OR under the
sceptical and literature-informed priors; the faint outer band is the prior-sensitivity range.
:::
::: {.it}
**Figura 1.** Odds ratio per studio (set primario, 11 studi con comparatori antibiotici).
Grigio = stime osservate; blu = stime a posteriori bayesiane (shrinkage), attratte verso la
media aggregata. La linea navy e la banda blu indicano l'OR aggregato (prior primario) e
l'ICr al 95%. Le linee verticali punteggiata e tratteggiata indicano l'OR aggregato con i
prior scettico e informato dalla letteratura; la banda esterna tenue è l'intervallo di
sensibilità ai prior.
:::
:::

## [Prior sensitivity]{.en}[Sensibilità ai prior]{.it}

::: {.en}
The estimate is stable across data-dominated priors (primary, vague, sceptical) but
shifts toward protection under the literature-informed prior centred on the earlier Tariq
(2018) estimate of OR 0.62 — i.e. that conclusion is driven by prior belief, not the new data.
:::
::: {.it}
La stima è stabile tra i prior dominati dai dati (primario, vago, scettico) ma si sposta
verso la protezione con il prior informato dalla letteratura, centrato sulla stima
precedente di Tariq (2018), OR 0,62 — cioè quella conclusione è guidata dalla convinzione
a priori, non dai nuovi dati.
:::

::: {.en}
```{r tbl-prior-en}
res_abx |>
  transmute(
    `Prior on μ` = c("Primary \u2014 N(0, 1\u00b2)", "Vague \u2014 N(0, 10\u00b2)",
                          "Sceptical \u2014 N(0, 0.35\u00b2)", "Literature \u2014 N(log 0.62, 0.10\u00b2)"),
    `OR (95% CrI)` = fmt_ci(OR, OR_lo, OR_hi),
    `P(OR<1)` = sprintf("%.2f", P_OR_lt_1),
    `P(OR<0.80)` = sprintf("%.2f", P_OR_lt_080)) |>
  kable(align = "lccc")
```
:::
::: {.it}
```{r tbl-prior-it}
res_abx |>
  transmute(
    `Prior su μ` = c("Primario \u2014 N(0, 1\u00b2)", "Vago \u2014 N(0, 10\u00b2)",
                          "Scettico \u2014 N(0, 0,35\u00b2)", "Letteratura \u2014 N(log 0,62, 0,10\u00b2)"),
    `OR (ICr 95%)` = fmt_ci(OR, OR_lo, OR_hi),
    `P(OR<1)` = sprintf("%.2f", P_OR_lt_1),
    `P(OR<0,80)` = sprintf("%.2f", P_OR_lt_080)) |>
  kable(align = "lccc")
```
:::

## [Prior-data conflict]{.en}[Conflitto prior-dati]{.it}

::: {.en}
Because the literature-informed prior is the only assumption under which the result becomes
confidently protective, we formally check whether it conflicts with the updated data. Using
the Box / Marshall-Spiegelhalter measure, we compare the prior for μ — N(log 0.62, 0.10²),
i.e. OR 0.62 — against the data-only estimate (vague-prior posterior:
OR `r sprintf("%.2f", exp(pdc_md))`, log-scale mean `r sprintf("%.2f", pdc_md)`,
SD `r sprintf("%.2f", pdc_sd)`). The conflict statistic is z = `r sprintf("%.2f", pdc_z)`
(two-sided p = `r sprintf("%.2f", pdc_p)`), indicating **mild tension but no formal
conflict**. The literature prior is more optimistic than the updated evidence, but the gap
is within sampling variation. Its influence on the pooled estimate comes mainly from being
far narrower (SD 0.10) than the data-only posterior (SD `r sprintf("%.2f", pdc_sd)`) rather
than from statistical incompatibility with the data.
:::
::: {.it}
Poiché il prior informato dalla letteratura è l'unica assunzione sotto cui il risultato
diventa decisamente protettivo, verifichiamo formalmente se sia in conflitto con i dati
aggiornati. Con la misura di Box / Marshall-Spiegelhalter confrontiamo il prior per μ —
N(log 0,62, 0,10²), cioè OR 0,62 — con la stima basata sui soli dati (posteriori con prior
vago: OR `r sprintf("%.2f", exp(pdc_md))`, media log `r sprintf("%.2f", pdc_md)`,
SD `r sprintf("%.2f", pdc_sd)`). La statistica di conflitto è z = `r sprintf("%.2f", pdc_z)`
(p bilaterale = `r sprintf("%.2f", pdc_p)`), indicando **una lieve tensione ma nessun
conflitto formale**. Il prior della letteratura è più ottimistico delle evidenze aggiornate,
ma la differenza rientra nella variabilità campionaria. La sua influenza deriva soprattutto
dall'essere molto più stretto (SD 0,10) della posteriori basata sui dati
(SD `r sprintf("%.2f", pdc_sd)`), non da un'incompatibilità statistica con i dati.
:::

## [MCMC confirmation & convergence]{.en}[Conferma MCMC e convergenza]{.it}

```{r brms-fit}
library(brms)
# brms/Stan MCMC fit (4 chains) on the primary set, precomputed and stored as RDS so that
# rendering never triggers Stan recompilation. Reproduce with the commented call below
# (re-run whenever prim_abx changes, then re-save the RDS).
bfit <- readRDS("brms_fit_primary.rds")
# bfit <- brm(bf(yi | se(sei) ~ 1 + (1 | study_id)), data = prim_abx,
#             prior = c(prior(normal(0, 1), class = Intercept),
#                       prior(normal(0, 0.5), class = sd)),
#             chains = 4, iter = 4000, warmup = 1000, cores = 4,
#             control = list(adapt_delta = 0.99), seed = 7291, refresh = 0)

library(posterior)
dm  <- as_draws_df(bfit)[["b_Intercept"]]
dt  <- as_draws_df(bfit)[["sd_study_id__Intercept"]]
brms_row <- tibble(source = "brms (MCMC)",
                   OR = exp(median(dm)), OR_lo = exp(quantile(dm,.025)), OR_hi = exp(quantile(dm,.975)),
                   tau = median(dt), P1 = mean(dm < 0), P08 = mean(dm < log(0.80)))
bmeta_row <- tibble(source = "bayesmeta (exact)",
                    OR = P$OR, OR_lo = P$OR_lo, OR_hi = P$OR_hi,
                    tau = P$tau, P1 = P$P_OR_lt_1, P08 = P$P_OR_lt_080)
dg <- summarise_draws(
  as_draws_df(bfit)[, c("b_Intercept","sd_study_id__Intercept")],
  "rhat","ess_bulk","ess_tail","mcse_mean")
# Sampler health: divergent transitions and max-treedepth hits
np      <- brms::nuts_params(bfit)
n_div   <- sum(np$Value[np$Parameter == "divergent__"])
n_maxtd <- sum(np$Value[np$Parameter == "treedepth__"] >= 10)
n_draws <- nrow(as_draws_df(bfit))
```

::: {.en}
**Sampling.** The confirmatory fit uses Hamiltonian Monte Carlo (Stan, via `brms`) with
the same priors as the exact model: 4 chains of 4,000 iterations (1,000 warmup), giving
`r format(n_draws, big.mark = ",")` post-warmup draws, with `adapt_delta = 0.99` to
handle the funnel-shaped geometry typical of hierarchical models with few studies. The
run produced **`r n_div` divergent transitions** and `r n_maxtd` iterations at the
maximum tree depth, so the sampler explored the posterior without numerical failures.

**Convergence.** R-hat compares within- and between-chain variance; both parameters are
essentially 1.00 (max `r sprintf("%.3f", max(dg$rhat))`, below the conventional 1.01
threshold), so the four chains have converged to the same target distribution. Bulk ESS
(`r round(dg$ess_bulk[1])` for μ, `r round(dg$ess_bulk[2])` for τ) and tail ESS
(`r round(dg$ess_tail[1])` / `r round(dg$ess_tail[2])`) are large, meaning both the
centre of the posterior and the extreme quantiles — on which the 95% credible interval
and P(OR<1) depend — rest on the equivalent of thousands of independent draws; the Monte
Carlo standard error of the pooled log-OR is only
`r sprintf("%.3f", dg$mcse_mean[1])`. The trace plots (Figure 2) confirm this visually:
the four chains overlap around stable levels, with no trends, drift or sticking.

**Agreement between engines.** `bayesmeta` computes the posterior by deterministic
numerical integration (no Monte Carlo error at all), whereas `brms`/Stan approximates it
by sampling. That the two agree closely (table below) shows the reported posterior is a
property of the model and the data rather than of the computational method — and it
validates the exact `bayesmeta` posteriors used throughout this report.
:::
::: {.it}
**Campionamento.** Il modello di conferma usa Hamiltonian Monte Carlo (Stan, tramite
`brms`) con gli stessi prior del modello esatto: 4 catene da 4.000 iterazioni (1.000 di
warmup), per un totale di `r format(n_draws, big.mark = ".")` draw post-warmup, con
`adapt_delta = 0,99` per gestire la geometria "a imbuto" tipica dei modelli gerarchici
con pochi studi. L'esecuzione ha prodotto **`r n_div` transizioni divergenti** e
`r n_maxtd` iterazioni alla profondità massima dell'albero: il campionatore ha esplorato
la posteriori senza problemi numerici.

**Convergenza.** R-hat confronta la varianza entro e tra le catene; entrambi i parametri
sono sostanzialmente a 1,00 (max `r sprintf("%.3f", max(dg$rhat))`, sotto la soglia
convenzionale di 1,01), quindi le quattro catene sono convergenti alla stessa
distribuzione obiettivo. Gli ESS bulk (`r round(dg$ess_bulk[1])` per μ,
`r round(dg$ess_bulk[2])` per τ) e tail (`r round(dg$ess_tail[1])` /
`r round(dg$ess_tail[2])`) sono ampi: sia il centro della posteriori sia i quantili
estremi — da cui dipendono l'ICr al 95% e P(OR<1) — poggiano sull'equivalente di
migliaia di draw indipendenti; l'errore standard Monte Carlo del log-OR aggregato è solo
`r sprintf("%.3f", dg$mcse_mean[1])`. I trace plot (Figura 2) lo confermano
visivamente: le quattro catene si sovrappongono attorno a livelli stabili, senza trend,
deriva o blocchi.

**Accordo tra i motori.** `bayesmeta` calcola la posteriori per integrazione numerica
deterministica (senza alcun errore Monte Carlo), mentre `brms`/Stan la approssima per
campionamento. Il fatto che i due concordino strettamente (tabella sotto) mostra che la
posteriori riportata è una proprietà del modello e dei dati, non del metodo
computazionale — e convalida le posteriori esatte di `bayesmeta` usate in tutto il
report.
:::

::: {.en}
```{r tbl-confirm-en}
bind_rows(bmeta_row, brms_row) |>
  transmute(Method = source, `OR (95% CrI)` = fmt_ci(OR, OR_lo, OR_hi),
            `τ` = sprintf("%.2f", tau),
            `P(OR<1)` = sprintf("%.2f", P1), `P(OR<0.80)` = sprintf("%.2f", P08)) |>
  kable(align = "lcccc")

dg |> transmute(Parameter = c("\u03bc (log-OR)","\u03c4 (heterogeneity SD)"),
                `R-hat` = sprintf("%.3f", rhat),
                `Bulk ESS` = round(ess_bulk), `Tail ESS` = round(ess_tail),
                `MCSE (mean)` = sprintf("%.3f", mcse_mean)) |>
  kable(align = "lcccc")
```
:::
::: {.it}
```{r tbl-confirm-it}
bind_rows(bmeta_row, brms_row) |>
  transmute(Metodo = source, `OR (ICr 95%)` = fmt_ci(OR, OR_lo, OR_hi),
            `τ` = sprintf("%.2f", tau),
            `P(OR<1)` = sprintf("%.2f", P1), `P(OR<0,80)` = sprintf("%.2f", P08)) |>
  kable(align = "lcccc")

dg |> transmute(Parametro = c("\u03bc (log-OR)","\u03c4 (SD eterogeneità)"),
                `R-hat` = sprintf("%.3f", rhat),
                `ESS bulk` = round(ess_bulk), `ESS tail` = round(ess_tail),
                `MCSE (media)` = sprintf("%.3f", mcse_mean)) |>
  kable(align = "lcccc")
```
:::

```{r fig-trace, fig.width=8, fig.height=4}
library(bayesplot)
mcmc_trace(bfit, pars = c("b_Intercept", "sd_study_id__Intercept"))
```

::: {.cap}
::: {.en}
**Figure 2.** MCMC trace plots for the four chains (pooled log-OR μ and heterogeneity SD τ). Overlapping chains fluctuating around stable levels — with no trends, drift or sticking — indicate good mixing and stationarity.
:::
::: {.it}
**Figura 2.** Trace plot MCMC per le quattro catene (log-OR aggregato μ e SD dell'eterogeneità τ). Catene sovrapposte che oscillano attorno a livelli stabili — senza trend, deriva o blocchi — indicano buon mixing e stazionarietà.
:::
:::

## [Small-study effects]{.en}[Effetti dei piccoli studi]{.it}

::: {.en}
With k = 11 studies (meeting the pre-specified ≥10 threshold), a contour-enhanced funnel
plot and Egger's regression test assess funnel asymmetry. There is **no evidence of
asymmetry** (Egger z = `r sprintf("%.2f", egg$zval)`, p = `r sprintf("%.2f", egg$pval)`);
the studies scatter fairly symmetrically about the pooled effect. Given the observational
evidence base and modest k, this is reassuring but not decisive.
:::
::: {.it}
Con k = 11 studi (soglia pre-specificata ≥10), un funnel plot contour-enhanced e il test di
regressione di Egger valutano l'asimmetria. **Non vi è evidenza di asimmetria**
(Egger z = `r sprintf("%.2f", egg$zval)`, p = `r sprintf("%.2f", egg$pval)`); gli studi si
distribuiscono in modo abbastanza simmetrico attorno all'effetto aggregato. Data la natura
osservazionale e il k contenuto, è rassicurante ma non decisivo.
:::

```{r fig-funnel, fig.width=7, fig.height=5.5}
funnel(rma_abx, level = c(90, 95, 99),
       shade = c("white", "#D7E0E8", "#B4C6D6"),
       refline = 0, col = jama_navy, bg = jama_blue,
       legend = TRUE, xlab = "Log odds ratio")
```

::: {.cap}
::: {.en}
**Figure 3.** Contour-enhanced funnel plot; shaded bands mark the p = 0.10 / 0.05 / 0.01 regions.
:::
::: {.it}
**Figura 3.** Funnel plot contour-enhanced; le bande indicano le regioni p = 0,10 / 0,05 / 0,01.
:::
:::

## [Leave-one-out]{.en}[Analisi leave-one-out]{.it}

::: {.en}
No single study drives the pooled estimate: removing any one study leaves the pooled OR
between `r sprintf("%.2f", min(loo$OR))` and `r sprintf("%.2f", max(loo$OR))`, always below
1. The most influential study is Saito 2025 (a harmful outlier); dropping it moves the OR
most toward protection, while dropping the protective Watson 2018 moves it toward the null.
The qualitative conclusion is stable throughout.
:::
::: {.it}
Nessun singolo studio determina la stima aggregata: rimuovendo un qualsiasi studio l'OR
aggregato resta tra `r sprintf("%.2f", min(loo$OR))` e `r sprintf("%.2f", max(loo$OR))`,
sempre inferiore a 1. Lo studio più influente è Saito 2025 (un outlier sfavorevole): la sua
rimozione sposta l'OR verso la protezione, mentre rimuovere il protettivo Watson 2018 lo
sposta verso l'assenza di effetto. La conclusione qualitativa resta stabile.
:::

```{r fig-loo, fig.width=8, fig.height=5}
lp <- getp("primary")
loo |> arrange(OR) |> mutate(dropped = factor(dropped, levels = dropped)) |>
  ggplot(aes(OR, dropped)) +
  annotate("rect", xmin = lp$OR_lo, xmax = lp$OR_hi, ymin = -Inf, ymax = Inf,
           fill = jama_band, alpha = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#4D4D4D", linewidth = 0.4) +
  geom_vline(xintercept = lp$OR, colour = jama_navy, linewidth = 0.7) +
  geom_pointrange(aes(xmin = OR_lo, xmax = OR_hi), colour = jama_gray, size = 0.4) +
  scale_x_log10(breaks = c(0.5, 0.7, 1, 1.3)) +
  labs(x = "Pooled OR after removing the study (log scale)", y = "Study removed") +
  theme_jama()
```

::: {.cap}
::: {.en}
**Figure 4.** Leave-one-out pooled OR; the blue line and band are the full k = 11 estimate and 95% CrI.
:::
::: {.it}
**Figura 4.** OR aggregato leave-one-out; la linea e la banda blu sono la stima completa k = 11 e l'ICr al 95%.
:::
:::

## [Doxycycline-only sensitivity analysis]{.en}[Analisi di sensibilità: sola doxiciclina]{.it}

```{r doxy-sens}
# Pre-specified sensitivity analysis: the primary antibiotic-comparator set (k = 11)
# restricted to doxycycline-exposure studies, keeping the primary prior. Reuses the
# doxycycline row already computed in `res_sub` (see setup).
D <- as.list(res_sub[res_sub$analysis == "doxycycline", ])
```

::: {.en}
Doxycycline is the dominant tetracycline in the evidence base and the agent most often
invoked for anaerobe-sparing activity, so we pre-specified a sensitivity analysis restricting
the primary antibiotic-comparator set to doxycycline-exposure studies (k = `r D$k`), keeping
the primary prior. The pooled odds ratio is **`r fmt_ci(D$OR, D$OR_lo, D$OR_hi)`** with
P(OR<1) = `r sprintf("%.2f", D$P_OR_lt_1)` and P(OR<0.80) = `r sprintf("%.2f", D$P_OR_lt_080)`. Restricting
to a single agent roughly halves the between-study heterogeneity
(τ = `r sprintf("%.2f", D$tau)` vs `r sprintf("%.2f", P$tau)` in the full primary set) and
tightens the predictive interval for a new study to OR
`r sprintf("%.2f\u2013%.2f", D$pred_lo, D$pred_hi)`. The doxycycline estimate is more precise
and more consistently below 1 than the all-tetracycline estimate, but its 95% credible
interval still touches 1 — a stronger, yet still not definitive, protective signal.
:::
::: {.it}
La doxiciclina è la tetraciclina prevalente nella base di evidenze e l'agente più spesso
citato per l'attività di risparmio degli anaerobi; abbiamo quindi pre-specificato un'analisi
di sensibilità che limita il set primario con comparatore antibiotico agli studi con
esposizione a doxiciclina (k = `r D$k`), mantenendo il prior primario. L'odds ratio aggregato
è **`r fmt_ci(D$OR, D$OR_lo, D$OR_hi)`** con P(OR<1) = `r sprintf("%.2f", D$P_OR_lt_1)` e
P(OR<0,80) = `r sprintf("%.2f", D$P_OR_lt_080)`. Limitarsi a un singolo agente dimezza circa
l'eterogeneità tra studi (τ = `r sprintf("%.2f", D$tau)` vs `r sprintf("%.2f", P$tau)` nel set
primario completo) e restringe l'intervallo predittivo per un nuovo studio a OR
`r sprintf("%.2f\u2013%.2f", D$pred_lo, D$pred_hi)`. La stima per la doxiciclina è più precisa
e più costantemente inferiore a 1 rispetto alla stima su tutte le tetracicline, ma il suo
intervallo di credibilità al 95% tocca ancora 1 — un segnale protettivo più forte, ma non
ancora definitivo.
:::

::: {.en}
```{r tbl-doxy-en}
tibble(
  Analysis = c("All tetracyclines (primary)", "Doxycycline only"),
  `Studies (k)` = c(nrow(prim_abx), D$k),
  `OR (95% CrI)` = c(fmt_ci(P$OR, P$OR_lo, P$OR_hi), fmt_ci(D$OR, D$OR_lo, D$OR_hi)),
  `τ` = sprintf("%.2f", c(P$tau, D$tau)),
  `P(OR<1)` = sprintf("%.2f", c(P$P_OR_lt_1, D$P_OR_lt_1)),
  `P(OR<0.80)` = sprintf("%.2f", c(P$P_OR_lt_080, D$P_OR_lt_080))
) |>
  kable(align = "lccccc")
```
:::
::: {.it}
```{r tbl-doxy-it}
tibble(
  Analisi = c("Tutte le tetracicline (primario)", "Solo doxiciclina"),
  `Studi (k)` = c(nrow(prim_abx), D$k),
  `OR (ICr 95%)` = c(fmt_ci(P$OR, P$OR_lo, P$OR_hi), fmt_ci(D$OR, D$OR_lo, D$OR_hi)),
  `τ` = sprintf("%.2f", c(P$tau, D$tau)),
  `P(OR<1)` = sprintf("%.2f", c(P$P_OR_lt_1, D$P_OR_lt_1)),
  `P(OR<0,80)` = sprintf("%.2f", c(P$P_OR_lt_080, D$P_OR_lt_080))
) |>
  kable(align = "lccccc")
```
:::

::: {.cap}
::: {.en}
**Table 2.** Doxycycline-only sensitivity analysis versus the full primary set. Restricting
to doxycycline exposure lowers heterogeneity and sharpens the protective signal, though the
credible interval still includes OR = 1.
:::
::: {.it}
**Tabella 2.** Analisi di sensibilità sulla sola doxiciclina rispetto al set primario
completo. Limitarsi all'esposizione a doxiciclina riduce l'eterogeneità e rende più netto il
segnale protettivo, benché l'intervallo di credibilità includa ancora OR = 1.
:::
:::

```{r fig-doxy, fig.width=7.5, fig.height=4}
doxy_studies <- prim_abx |>
  dplyr::filter(exposure_class == "doxycycline") |>
  transmute(label = reference, type = "Doxycycline study (observed)",
            OR = exp(yi), OR_lo = exp(yi - 1.96 * sei), OR_hi = exp(yi + 1.96 * sei))
pooled_rows <- tibble(
  label = c("Doxycycline only (pooled, k = 6)", "All tetracyclines (pooled, k = 11)"),
  type  = "Pooled estimate",
  OR    = c(D$OR, P$OR), OR_lo = c(D$OR_lo, P$OR_lo), OR_hi = c(D$OR_hi, P$OR_hi))
lev <- c("All tetracyclines (pooled, k = 11)", "Doxycycline only (pooled, k = 6)",
         doxy_studies |> arrange(desc(OR)) |> pull(label))
bind_rows(doxy_studies, pooled_rows) |>
  mutate(label = factor(label, levels = lev),
         type  = factor(type, levels = c("Doxycycline study (observed)", "Pooled estimate"))) |>
  ggplot(aes(OR, label, colour = type)) +
  annotate("rect", xmin = D$OR_lo, xmax = D$OR_hi, ymin = -Inf, ymax = Inf,
           fill = jama_band, alpha = 0.4) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#4D4D4D", linewidth = 0.4) +
  geom_pointrange(aes(xmin = OR_lo, xmax = OR_hi), size = 0.45) +
  geom_text(aes(label = sprintf("%.2f (%.2f-%.2f)", OR, OR_lo, OR_hi)),
            vjust = -1, size = 3.2, colour = "#1A1A1A") +
  scale_colour_manual(values = c("Doxycycline study (observed)" = jama_gray,
                                 "Pooled estimate" = jama_navy)) +
  scale_x_log10(breaks = c(0.1, 0.25, 0.5, 1, 2, 4)) +
  labs(x = "Odds ratio (log scale)", y = NULL, colour = NULL) +
  theme_jama()
```

::: {.cap}
::: {.en}
**Figure 5.** Forest plot of the doxycycline-only sensitivity analysis. Gray points are the
six doxycycline-exposure studies (observed ORs, 95% CIs); navy points are the pooled Bayesian
estimates for the doxycycline-only subset (k = 6) and the full primary set (k = 11). The
shaded band is the doxycycline-only 95% credible interval; the dashed line marks OR = 1.
:::
::: {.it}
**Figura 5.** Forest plot dell'analisi di sensibilità sulla sola doxiciclina. I punti grigi
sono i sei studi con esposizione a doxiciclina (OR osservati, IC 95%); i punti navy sono le
stime bayesiane aggregate per il sottoinsieme doxiciclina (k = 6) e per il set primario
completo (k = 11). La banda ombreggiata è l'intervallo di credibilità al 95% della sola
doxiciclina; la linea tratteggiata indica OR = 1.
:::
:::

## [Additional sensitivity analyses]{.en}[Analisi di sensibilità aggiuntive]{.it}

::: {.en}
**Adjusted-only secondary comparison.** Restricting the no-antibiotic comparison to the
three studies with multivariable-adjusted estimates (Delaney 2007, Dial 2008, Brown 2016)
gives OR `r fmt_ci(sens_sec_adj$OR, sens_sec_adj$OR_lo, sens_sec_adj$OR_hi)` (k = 3),
P(OR<1) = `r sprintf("%.2f", sens_sec_adj$P_OR_lt_1)` — consistent with the full
secondary set, with a wider interval reflecting the smaller evidence base.

**Excluding ratio-metric conversions.** The curated estimates of Doernberg 2012 (adjusted
HR), Dial 2008 (adjusted RR) and Brown 2016 (adjusted IRR) enter the analysis as ORs
under a rare-outcome approximation. Excluding them leaves the primary estimate
essentially unchanged (k = 10; OR
`r fmt_ci(sens_prim_or$OR, sens_prim_or$OR_lo, sens_prim_or$OR_hi)`, P(OR<1) =
`r sprintf("%.2f", sens_prim_or$P_OR_lt_1)`) and the secondary estimate likewise
(k = 7; OR `r fmt_ci(sens_sec_or$OR, sens_sec_or$OR_lo, sens_sec_or$OR_hi)`, P(OR<1) =
`r sprintf("%.2f", sens_sec_or$P_OR_lt_1)`). Conclusions are robust to this
approximation.
:::
::: {.it}
**Confronto secondario con sole stime aggiustate.** Limitando il confronto contro nessun
antibiotico ai tre studi con stime aggiustate multivariabili (Delaney 2007, Dial 2008,
Brown 2016) si ottiene OR
`r fmt_ci(sens_sec_adj$OR, sens_sec_adj$OR_lo, sens_sec_adj$OR_hi)` (k = 3),
P(OR<1) = `r sprintf("%.2f", sens_sec_adj$P_OR_lt_1)` — coerente con l'intero set
secondario, con un intervallo più ampio che riflette la base di evidenze ridotta.

**Esclusione delle conversioni da misure di tasso/rischio.** Le stime curate di
Doernberg 2012 (HR aggiustato), Dial 2008 (RR aggiustato) e Brown 2016 (IRR aggiustato)
entrano nell'analisi come OR tramite un'approssimazione per esiti rari. Escludendole, la
stima primaria resta sostanzialmente invariata (k = 10; OR
`r fmt_ci(sens_prim_or$OR, sens_prim_or$OR_lo, sens_prim_or$OR_hi)`, P(OR<1) =
`r sprintf("%.2f", sens_prim_or$P_OR_lt_1)`), così come quella secondaria (k = 7; OR
`r fmt_ci(sens_sec_or$OR, sens_sec_or$OR_lo, sens_sec_or$OR_hi)`, P(OR<1) =
`r sprintf("%.2f", sens_sec_or$P_OR_lt_1)`). Le conclusioni sono robuste rispetto a
questa approssimazione.
:::

## [Secondary comparison: vs no antibiotic (9 studies)]{.en}[Confronto secondario: vs nessun antibiotico (9 studi)]{.it}

::: {.en}
Against a *no-antibiotic* comparator the association is essentially null
(**OR `r fmt_ci(sec$OR, sec$OR_lo, sec$OR_hi)`**, P(OR<1) = `r sprintf("%.2f", sec$P_OR_lt_1)`).
This is expected — almost any antibiotic raises CDI risk relative to no antibiotic — and it
explains why pooling these studies into the primary set attenuated the protective signal.
The two comparator questions are distinct and should not be combined.
:::
::: {.it}
Rispetto a un comparatore *senza antibiotico* l'associazione è sostanzialmente nulla
(**OR `r fmt_ci(sec$OR, sec$OR_lo, sec$OR_hi)`**, P(OR<1) = `r sprintf("%.2f", sec$P_OR_lt_1)`).
È un risultato atteso — quasi ogni antibiotico aumenta il rischio di ICD rispetto a nessun
antibiotico — e spiega perché aggregare questi studi nel set primario attenuava il segnale
protettivo. Le due domande sul comparatore sono distinte e non vanno combinate.
:::

## [Subgroups]{.en}[Sottogruppi]{.it}

::: {.en}
Within the primary set, the doxycycline-restricted subgroup and cohort studies lean
protective, whereas case-control studies do not — a design-related divergence that warrants
cautious interpretation.
:::
::: {.it}
All'interno del set primario, il sottogruppo limitato alla doxiciclina e gli studi di
coorte tendono alla protezione, mentre gli studi caso-controllo no: una divergenza legata
al disegno da interpretare con cautela.
:::

::: {.en}
```{r tbl-sub-en}
res_sub |>
  transmute(Analysis = recode(analysis, doxycycline = "Doxycycline only",
              case_control = "Case-control designs", cohort = "Cohort designs"),
            `Studies (k)` = k, `OR (95% CrI)` = fmt_ci(OR, OR_lo, OR_hi),
            `P(OR<1)` = sprintf("%.2f", P_OR_lt_1)) |>
  kable(align = "lccc")
```
:::
::: {.it}
```{r tbl-sub-it}
res_sub |>
  transmute(Analisi = recode(analysis, doxycycline = "Solo doxiciclina",
              case_control = "Disegni caso-controllo", cohort = "Disegni di coorte"),
            `Studi (k)` = k, `OR (ICr 95%)` = fmt_ci(OR, OR_lo, OR_hi),
            `P(OR<1)` = sprintf("%.2f", P_OR_lt_1)) |>
  kable(align = "lccc")
```
:::

## [Care-context meta-regression]{.en}[Meta-regressione per contesto assistenziale]{.it}

::: {.en}
A pre-specified Bayesian meta-regression (exact, `bmr`) on the **primary antibiotic-comparator
set (k = 11)** estimates a separate pooled OR per care context (CA 3, HA 4, mixed 4). The
protective association is concentrated in **hospital-acquired (inpatient) settings**
(OR `r fmt_ci(ctx$OR[2], ctx$OR_lo[2], ctx$OR_hi[2])`,
P(OR<1) = `r sprintf("%.2f", ctx$P_OR_lt_1[2])`), while community-acquired and mixed settings
straddle 1. This mirrors Tariq (2018), whose setting subgroup found protection in inpatients
(OR 0.55) but not mixed populations (OR 0.92). As a **sensitivity analysis on the full primary
set (k = 20; CA 6, HA 9, mixed 5)**, the hospital-acquired estimate attenuates to OR `r ha20`
(P(OR<1) = `r ha20_p`), because that set adds no-antibiotic-comparator studies that dilute the
setting contrast. Per-group counts are small, so this is exploratory; residual heterogeneity
remains (τ = `r mr_tau`).
:::
::: {.it}
Una meta-regressione bayesiana pre-specificata (esatta, `bmr`) sul **set primario con
comparatore antibiotico (k = 11)** stima un OR aggregato per ciascun contesto (CA 3, HA 4,
misto 4). L'associazione protettiva si concentra nei **contesti ospedalieri (pazienti
ricoverati)** (OR `r fmt_ci(ctx$OR[2], ctx$OR_lo[2], ctx$OR_hi[2])`,
P(OR<1) = `r sprintf("%.2f", ctx$P_OR_lt_1[2])`), mentre i contesti comunitari e misti
attraversano 1. Ciò rispecchia Tariq (2018), la cui analisi per contesto trovava protezione
nei ricoverati (OR 0,55) ma non nelle popolazioni miste (OR 0,92). Come **analisi di
sensibilità sul set primario completo (k = 20; CA 6, HA 9, misto 5)**, la stima ospedaliera si
attenua a OR `r ha20` (P(OR<1) = `r ha20_p`), perché quel set aggiunge studi con comparatore
"nessun antibiotico" che diluiscono il contrasto. Le numerosità per gruppo sono piccole,
quindi è esplorativo; l'eterogeneità residua permane (τ = `r mr_tau`).
:::

```{r fig-metareg, fig.width=7.5, fig.height=3.4}
bind_rows(ctx |> mutate(set = "Primary (k=11)"),
          ctx20 |> mutate(set = "Sensitivity (k=20)")) |>
  mutate(care_context = factor(care_context, levels = c("mixed","HA","CA")),
         set = factor(set, levels = c("Primary (k=11)","Sensitivity (k=20)"))) |>
  ggplot(aes(OR, care_context, colour = set)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#4D4D4D", linewidth = 0.4) +
  geom_pointrange(aes(xmin = OR_lo, xmax = OR_hi),
                  position = position_dodge(width = 0.5), size = 0.45) +
  scale_colour_manual(values = c("Primary (k=11)" = jama_blue,
                                 "Sensitivity (k=20)" = jama_gray)) +
  scale_x_log10(breaks = c(0.3, 0.5, 0.8, 1, 1.5, 2)) +
  labs(x = "Odds ratio (log scale)", y = "Care context") +
  theme_jama()
```

::: {.cap}
::: {.en}
**Figure 6.** Care-context pooled ORs from the Bayesian meta-regression: primary antibiotic-comparator set (k = 11, blue) with full-set sensitivity (k = 20, gray).
:::
::: {.it}
**Figura 6.** OR aggregati per contesto dalla meta-regressione bayesiana: set primario con comparatore antibiotico (k = 11, blu) e sensibilità sul set completo (k = 20, grigio).
:::
:::

## [Design & adjustment meta-regression]{.en}[Meta-regressione per disegno e aggiustamento]{.it}

::: {.en}
Two further pre-specified Bayesian meta-regressions (exact, `bmr`) complete the covariate set
named in the protocol (§19: context, design, adjustment). **Study design** is modelled on the
primary antibiotic-comparator set (k = 11), collapsing designs to case-control (k = 6) versus
cohort (k = 5). Cohort studies lean protective
(OR `r fmt_ci(dsn$OR[2], dsn$OR_lo[2], dsn$OR_hi[2])`,
P(OR<1) = `r sprintf("%.2f", dsn$P_OR_lt_1[2])`), whereas case-control studies straddle 1
(OR `r fmt_ci(dsn$OR[1], dsn$OR_lo[1], dsn$OR_hi[1])`,
P(OR<1) = `r sprintf("%.2f", dsn$P_OR_lt_1[1])`) — the same design-related divergence flagged
in the subgroup table. **Adjustment status** cannot be regressed within the primary set
because all 11 estimates are covariate-adjusted; it is therefore assessed on the full primary
set (k = 20; 14 adjusted, 6 unadjusted). Adjusted estimates are modestly protective
(OR `r fmt_ci(adjt$OR[1], adjt$OR_lo[1], adjt$OR_hi[1])`) while unadjusted estimates are
null-to-slightly-adverse (OR `r fmt_ci(adjt$OR[2], adjt$OR_lo[2], adjt$OR_hi[2])`), with
heavily overlapping intervals that both cross 1 — so there is no sign that unadjusted studies
inflate the protective signal (if anything, the reverse). Per-group counts are small and
intervals wide, so both analyses are exploratory; residual heterogeneity persists
(design τ = `r mrd_tau`; adjustment τ = `r mra_tau`).
:::
::: {.it}
Due ulteriori meta-regressioni bayesiane pre-specificate (esatte, `bmr`) completano l'insieme
di covariate indicato nel protocollo (§19: contesto, disegno, aggiustamento). Il **disegno
dello studio** è modellato sul set primario con comparatore antibiotico (k = 11), raggruppando
i disegni in caso-controllo (k = 6) vs coorte (k = 5). Gli studi di coorte tendono alla
protezione (OR `r fmt_ci(dsn$OR[2], dsn$OR_lo[2], dsn$OR_hi[2])`,
P(OR<1) = `r sprintf("%.2f", dsn$P_OR_lt_1[2])`), mentre i caso-controllo attraversano 1
(OR `r fmt_ci(dsn$OR[1], dsn$OR_lo[1], dsn$OR_hi[1])`,
P(OR<1) = `r sprintf("%.2f", dsn$P_OR_lt_1[1])`) — la stessa divergenza legata al disegno
evidenziata nella tabella dei sottogruppi. Lo **stato di aggiustamento** non può essere
regredito nel set primario perché tutte le 11 stime sono aggiustate; viene quindi valutato sul
set primario completo (k = 20; 14 aggiustate, 6 non aggiustate). Le stime aggiustate sono
modestamente protettive (OR `r fmt_ci(adjt$OR[1], adjt$OR_lo[1], adjt$OR_hi[1])`) mentre quelle
non aggiustate sono nulle-o-lievemente-avverse (OR `r fmt_ci(adjt$OR[2], adjt$OR_lo[2], adjt$OR_hi[2])`),
con intervalli ampiamente sovrapposti che attraversano entrambi 1 — quindi non vi è segno che
gli studi non aggiustati amplifichino il segnale protettivo (semmai il contrario). Le
numerosità per gruppo sono piccole e gli intervalli ampi, quindi entrambe le analisi sono
esplorative; l'eterogeneità residua permane (τ disegno = `r mrd_tau`; τ aggiustamento = `r mra_tau`).
:::

```{r fig-metareg2, fig.width=7.5, fig.height=3.4}
bind_rows(dsn  |> mutate(panel = "Study design (k = 11)"),
          adjt |> mutate(panel = "Adjustment status (k = 20)")) |>
  mutate(panel = factor(panel, levels = c("Study design (k = 11)", "Adjustment status (k = 20)")),
         group = factor(group, levels = c("Cohort", "Case-control", "Unadjusted", "Adjusted"))) |>
  ggplot(aes(OR, group)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#4D4D4D", linewidth = 0.4) +
  geom_pointrange(aes(xmin = OR_lo, xmax = OR_hi), colour = jama_blue, size = 0.45) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  scale_x_log10(breaks = c(0.3, 0.5, 0.8, 1, 1.5, 2)) +
  labs(x = "Odds ratio (log scale)", y = NULL) +
  theme_jama() +
  theme(strip.background = element_blank())
```

::: {.cap}
::: {.en}
**Figure 7.** Pooled odds ratios from the design and adjustment Bayesian meta-regressions (exact `bmr`, cell-means coding): study design on the primary antibiotic-comparator set (k = 11), adjustment on the full primary set (k = 20). Dashed line at OR = 1; all credible intervals cross it.
:::
::: {.it}
**Figura 7.** OR aggregati dalle meta-regressioni bayesiane per disegno e aggiustamento (`bmr` esatto, codifica cell-means): disegno sul set primario con comparatore antibiotico (k = 11), aggiustamento sul set primario completo (k = 20). Linea tratteggiata a OR = 1; tutti gli intervalli di credibilità la attraversano.
:::
:::

::: {.en}
```{r tbl-mr2-en}
bind_rows(dsn  |> mutate(Covariate = "Study design (k = 11)"),
          adjt |> mutate(Covariate = "Adjustment (k = 20)")) |>
  transmute(Covariate, Group = group, `Studies (k)` = k,
            `OR (95% CrI)` = fmt_ci(OR, OR_lo, OR_hi),
            `P(OR<1)` = sprintf("%.2f", P_OR_lt_1)) |>
  kable(align = "llccc")
```
:::
::: {.it}
```{r tbl-mr2-it}
bind_rows(dsn  |> mutate(Covariata = "Disegno (k = 11)"),
          adjt |> mutate(Covariata = "Aggiustamento (k = 20)")) |>
  transmute(Covariata, Gruppo = group, `Studi (k)` = k,
            `OR (ICr 95%)` = fmt_ci(OR, OR_lo, OR_hi),
            `P(OR<1)` = sprintf("%.2f", P_OR_lt_1)) |>
  kable(align = "llccc")
```
:::

::: {.cap}
::: {.en}
**Design and adjustment meta-regression.** Pooled ORs per group (exact Bayesian `bmr`, cell-means coding). Design is modelled on the primary antibiotic-comparator set (k = 11); adjustment on the full primary set (k = 20), because all 11 primary-comparator estimates are adjusted. Both are exploratory given small per-group counts.
:::
::: {.it}
**Meta-regressione per disegno e aggiustamento.** OR aggregati per gruppo (`bmr` bayesiano esatto, codifica cell-means). Il disegno è modellato sul set primario con comparatore antibiotico (k = 11); l'aggiustamento sul set primario completo (k = 20), poiché tutte le 11 stime con comparatore antibiotico sono aggiustate. Entrambe sono esplorative viste le piccole numerosità per gruppo.
:::
:::

## [Comparison with Tariq (2018)]{.en}[Confronto con Tariq (2018)]{.it}

::: {.en}
**Can our Bayesian approach reproduce Tariq?** Yes. Applied to Tariq's exact six studies,
a DerSimonian-Laird model reproduces their result exactly (`r v_tariq_dl`, I² = 53%), and
our Bayesian model gives a near-identical point estimate (`r v_tariq_bay`,
P(OR<1) = `r v_tariq_p`). The Bayesian credible interval is wider than Tariq's confidence
interval because it propagates the uncertainty in τ that is unavoidable with only six
studies — the divergence is one of honest uncertainty, not of the point estimate.

**Why does the updated review not find clear protection?** Chiefly the **additional
studies**. Adding the eight newer antibiotic-comparator studies to Tariq's six moves the
pooled OR from ~0.63 to `r v_bridge` (P(OR<1) = `r v_bridge_p`) and roughly doubles
heterogeneity — and this bridge lands essentially on our primary estimate. The new evidence
is dominated by very large administrative cohorts (e.g. Carmichael, Miller, Watson, Webb,
Boven) whose estimates sit close to the null and carry substantial weight. Secondary
contributors to the difference are (i) the **method** (Bayesian vs DL), which mainly widens
intervals rather than shifting the centre; and (ii) **comparator and estimate decisions** —
our protocol separates "vs other antibiotics" from "vs no antibiotic", so three of Tariq's
six studies (Delaney, Dial, Tartof) move to the secondary set, and Tartof's estimate
differs (Tariq's adjusted OR 0.50 vs our unadjusted 0.84).
:::
::: {.it}
**Il nostro approccio bayesiano riproduce Tariq?** Sì. Applicato ai sei studi esatti di
Tariq, un modello DerSimonian-Laird ne riproduce esattamente il risultato (`r v_tariq_dl`,
I² = 53%), e il nostro modello bayesiano fornisce una stima puntuale quasi identica
(`r v_tariq_bay`, P(OR<1) = `r v_tariq_p`). L'intervallo di credibilità bayesiano è più
ampio dell'intervallo di confidenza di Tariq perché propaga l'incertezza su τ, inevitabile
con soli sei studi: la differenza riguarda l'incertezza, non la stima puntuale.

**Perché la revisione aggiornata non trova una protezione netta?** Soprattutto per gli
**studi aggiuntivi**. Aggiungendo gli otto nuovi studi con comparatore antibiotico ai sei
di Tariq, l'OR aggregato passa da ~0,63 a `r v_bridge` (P(OR<1) = `r v_bridge_p`) e
l'eterogeneità circa raddoppia — e questo "ponte" coincide sostanzialmente con la nostra
stima primaria. Le nuove evidenze sono dominate da coorti amministrative molto grandi (es.
Carmichael, Miller, Watson, Webb, Boven), con stime vicine al valore nullo e peso
sostanziale. Contribuiscono in via secondaria (i) il **metodo** (bayesiano vs DL), che
allarga gli intervalli più che spostare il centro; e (ii) le **decisioni su comparatore e
stime** — il protocollo separa "vs altri antibiotici" da "vs nessun antibiotico", per cui
tre dei sei studi di Tariq (Delaney, Dial, Tartof) passano al set secondario, e la stima di
Tartof differisce (OR aggiustato 0,50 di Tariq vs 0,84 non aggiustato nostro).
:::

```{r fig-tariq, fig.width=8, fig.height=3.2}
comp |> mutate(analysis = factor(analysis, levels = rev(analysis))) |>
  ggplot(aes(OR, analysis)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#4D4D4D", linewidth = 0.4) +
  geom_pointrange(aes(xmin = OR_lo, xmax = OR_hi), colour = jama_navy, size = 0.5) +
  geom_text(aes(label = sprintf("%.2f (%.2f-%.2f)", OR, OR_lo, OR_hi)),
            vjust = -1, size = 3.2, colour = "#1A1A1A") +
  scale_x_log10(breaks = c(0.4, 0.5, 0.6, 0.8, 1, 1.2)) +
  labs(x = "Pooled OR (log scale)", y = NULL) +
  theme_jama()
```

::: {.cap}
::: {.en}
**Figure 8.** From Tariq to the updated review: recapitulation and stepwise decomposition of the difference.
:::
::: {.it}
**Figura 8.** Da Tariq alla revisione aggiornata: recupero e decomposizione passo-passo della differenza.
:::
:::

## [Risk-of-bias summary]{.en}[Sintesi del rischio di bias]{.it}

::: {.en}
Figure 9 summarises the study-level Newcastle-Ottawa judgements as a traffic-light plot,
mapping each study's per-domain star counts to Low / Some concerns / High. All 20 studies are
**Low** overall except Predrag 2016 (some concerns); the scattered amber cells sit almost
entirely in the Selection domain (representativeness / non-response), while the comparability
and outcome/exposure domains are largely green.
:::
::: {.it}
La Figura 9 riassume le valutazioni Newcastle-Ottawa per studio come traffic-light plot,
mappando i conteggi di stelle per dominio in Basso / Alcune perplessità / Alto. Tutti i 20
studi sono complessivamente a rischio **basso** tranne Predrag 2016 (alcune perplessità); le
celle ambra si concentrano quasi esclusivamente nel dominio Selezione (rappresentatività /
non-risposta), mentre i domini di comparabilità ed esito/esposizione sono in gran parte verdi.
:::

```{r fig-rob, fig.width=8, fig.height=6}
rob_cols <- c(Low = "#2E7D32", `Some concerns` = "#F4B400", High = "#C62828")
sym      <- c(Low = "+", `Some concerns` = "-", High = "x")
tl <- readr::read_csv("nos_assessment.csv", show_col_types = FALSE) |>
  transmute(study, total,
    Selection = sel1 + sel2 + sel3 + sel4,
    Comparability = comp1 + comp2,
    `Outcome/Exposure` = oe1 + oe2 + oe3,
    Overall = stringr::str_to_sentence(rob_overall)) |>
  mutate(
    Selection = dplyr::case_when(Selection >= 4 ~ "Low", Selection >= 2 ~ "Some concerns", TRUE ~ "High"),
    Comparability = dplyr::case_when(Comparability >= 2 ~ "Low", Comparability == 1 ~ "Some concerns", TRUE ~ "High"),
    `Outcome/Exposure` = dplyr::case_when(`Outcome/Exposure` >= 3 ~ "Low", `Outcome/Exposure` == 2 ~ "Some concerns", TRUE ~ "High")) |>
  dplyr::arrange(total, dplyr::desc(study))
tl_long <- tl |>
  tidyr::pivot_longer(c(Selection, Comparability, `Outcome/Exposure`, Overall),
                      names_to = "Domain", values_to = "Judgement") |>
  mutate(study = factor(study, levels = tl$study),
         Domain = factor(Domain, levels = c("Selection","Comparability","Outcome/Exposure","Overall")),
         Judgement = factor(Judgement, levels = c("Low","Some concerns","High")))
ggplot(tl_long, aes(Domain, study)) +
  geom_point(aes(colour = Judgement), size = 7) +
  geom_text(aes(label = sym[as.character(Judgement)]), colour = "white", size = 3.6, fontface = "bold") +
  geom_vline(xintercept = 3.5, linetype = "dotted", colour = "#4D4D4D") +
  scale_colour_manual(values = rob_cols, drop = FALSE) +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL, colour = "Risk of bias") +
  theme_jama() +
  theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 20, hjust = 0))
```

::: {.cap}
::: {.en}
**Figure 9.** Risk-of-bias traffic-light plot (Newcastle-Ottawa Scale). Per-domain judgements are derived from item stars — Selection (max 4), Comparability (max 2), Outcome/Exposure (max 3) — and Overall is the study-level NOS rating. Green **+** = low, amber **−** = some concerns, red **×** = high. Provisional, single-reviewer assessment.
:::
::: {.it}
**Figura 9.** Traffic-light plot del rischio di bias (scala Newcastle-Ottawa). I giudizi per dominio derivano dalle stelle per item — Selezione (max 4), Comparabilità (max 2), Esito/Esposizione (max 3) — e il giudizio complessivo è la valutazione NOS per studio. Verde **+** = basso, ambra **−** = alcune perplessità, rosso **×** = alto. Valutazione provvisoria, singolo revisore.
:::
:::

## [Certainty of evidence (GRADE)]{.en}[Certezza delle evidenze (GRADE)]{.it}

::: {.en}
Certainty was rated following the GRADE approach [@Guyatt2008GRADE]. Because all evidence
is observational, both comparisons start at *low* certainty; each was rated down for
serious inconsistency and serious imprecision, yielding **very low** certainty for both
questions.
:::

::: {.en}
| Outcome: incident CDI | k | Pooled OR (95% CrI) | Illustrative absolute effect* | Risk of bias | Inconsistency | Indirectness | Imprecision | Publication bias | Certainty |
|---|---|---|---|---|---|---|---|---|---|
| Tetracyclines vs other antibiotics | 11 | `r fmt_ci(P$OR, P$OR_lo, P$OR_hi)` | `r abs_prim` | Not seriousᵃ | Seriousᵇ | Not serious | Seriousᶜ | Undetectedᵈ | **Very low** |
| Tetracyclines vs no antibiotic | 9 | `r fmt_ci(sec$OR, sec$OR_lo, sec$OR_hi)` | `r abs_sec` | Not seriousᵃ | Seriousᵇ | Seriousᵉ | Seriousᶜ | Undetectedᵈ | **Very low** |

\* Illustrative only: assumes a baseline risk of 20 per 1,000 antibiotic-exposed patients
(hospital context) and 5 per 1,000 unexposed patients; absolute effects scale with the
baseline risk. ᵃ All 11 primary and 8 of 9 secondary studies at low risk of bias on the
Newcastle-Ottawa Scale (provisional, single-reviewer assessment). ᵇ Substantial
heterogeneity: primary I² = `r i2_txt`; the predictive interval spans protection to harm.
ᶜ Credible interval includes both meaningful benefit and harm. ᵈ No funnel-plot asymmetry
(Egger's test p = `r sprintf("%.2f", egg$pval)`), though power to detect it is limited.
ᵉ The no-antibiotic comparator answers a different question and is highly prone to
confounding by indication.
:::

::: {.it}
La certezza è stata valutata secondo l'approccio GRADE [@Guyatt2008GRADE]. Poiché tutte
le evidenze sono osservazionali, entrambi i confronti partono da una certezza *bassa*;
ciascuno è stato declassato per inconsistenza e imprecisione serie, risultando in una
certezza **molto bassa** per entrambe le domande.
:::

::: {.it}
| Esito: ICD incidente | k | OR aggregato (ICr 95%) | Effetto assoluto illustrativo* | Rischio di bias | Inconsistenza | Indirettezza | Imprecisione | Bias di pubblicazione | Certezza |
|---|---|---|---|---|---|---|---|---|---|
| Tetracicline vs altri antibiotici | 11 | `r fmt_ci(P$OR, P$OR_lo, P$OR_hi)` | `r abs_prim` | Non serioᵃ | Serioᵇ | Non serio | Serioᶜ | Non rilevatoᵈ | **Molto bassa** |
| Tetracicline vs nessun antibiotico | 9 | `r fmt_ci(sec$OR, sec$OR_lo, sec$OR_hi)` | `r abs_sec` | Non serioᵃ | Serioᵇ | Serioᵉ | Serioᶜ | Non rilevatoᵈ | **Molto bassa** |

\* Solo illustrativo: assume un rischio di base di 20 per 1.000 pazienti esposti ad
antibiotici (contesto ospedaliero) e di 5 per 1.000 non esposti; gli effetti assoluti
variano con il rischio di base. ᵃ Tutti gli 11 studi primari e 8 dei 9 secondari a basso
rischio di bias secondo la scala Newcastle-Ottawa (valutazione provvisoria, singolo
revisore). ᵇ Eterogeneità sostanziale: I² primario = `r i2_txt`; l'intervallo predittivo
spazia dalla protezione al danno. ᶜ L'intervallo di credibilità include sia un beneficio
sia un danno rilevanti. ᵈ Nessuna asimmetria del funnel plot (test di Egger p =
`r sprintf("%.2f", egg$pval)`), sebbene la potenza per rilevarla sia limitata. ᵉ Il
comparatore senza antibiotico risponde a una domanda diversa ed è fortemente soggetto a
confondimento per indicazione.
:::

## [Interpretation]{.en}[Interpretazione]{.it}

::: {.en}
Restricted to the protocol's intended comparator (another antibiotic), the data provide
**suggestive but uncertain** evidence that tetracycline exposure lowers CDI risk. The
pooled odds ratio is below 1 (OR `r fmt_ci(P$OR, P$OR_lo, P$OR_hi)`) but its 95% credible
interval still includes no effect; the posterior probability of *any* reduction is
`r sprintf("%.0f", 100*P$P_OR_lt_1)`%, falling to `r sprintf("%.0f", 100*P$P_OR_lt_080)`%
for a clinically meaningful one (OR < 0.80). These findings distinguish a plausibly
*lower associated risk* from established *protective therapy*.

**What the model fit tells us.** The random-effects model describes the studies well, but
its most informative output is the between-study spread rather than the mean. Heterogeneity
is substantial ($\tau$ = `r sprintf("%.2f", P$tau)`, I² = `r i2_txt`), so the pooled OR
summarises a *distribution* of study-specific effects rather than one common value: the
predictive interval for a new study (OR `r sprintf("%.2f\u2013%.2f", P$pred_lo, P$pred_hi)`)
reaches from meaningful protection to modest harm. The partial pooling visible in Figure 1
shrinks imprecise study estimates toward that distribution, which is why the pooled interval
is wider than a fixed-effect analysis would imply. The exact numerical posterior was
reproduced by independent Hamiltonian Monte Carlo (R-hat ≈ 1.00, effective sample sizes in
the thousands, `r n_div` divergent transitions), so this width faithfully reflects the
evidence rather than any artefact of computation or prior choice.

**How robust is the result?** The direction of effect is stable across the pre-specified
checks: no single study drives it (leave-one-out OR
`r sprintf("%.2f", min(loo$OR))`–`r sprintf("%.2f", max(loo$OR))`), it persists after
excluding the three RR/HR/IRR-to-OR conversions (OR
`r fmt_ci(sens_prim_or$OR, sens_prim_or$OR_lo, sens_prim_or$OR_hi)`) and after restricting
to low-risk-of-bias studies, and there is no funnel asymmetry (Egger p =
`r sprintf("%.2f", egg$pval)`). It is likewise stable across the data-dominated priors
(primary, vague, sceptical). The only assumption that renders the result confidently
protective is the literature-informed prior centred on Tariq's OR 0.62; the formal
prior-data conflict check finds only mild tension (z = `r sprintf("%.2f", pdc_z)`, p =
`r sprintf("%.2f", pdc_p)`), so that shift reflects the prior's narrowness rather than
incompatibility with the updated data. The strongest internal signals — the doxycycline-only
subset (OR `r fmt_ci(D$OR, D$OR_lo, D$OR_hi)`) and hospital-acquired settings — are
consistent with a real but small effect concentrated in specific agents and contexts, yet
each interval still touches 1.
:::
::: {.it}
Limitandosi al comparatore previsto dal protocollo (un altro antibiotico), i dati
forniscono un'evidenza **suggestiva ma incerta** che l'esposizione alle tetracicline
riduca il rischio di ICD. L'odds ratio aggregato è inferiore a 1 (OR
`r fmt_ci(P$OR, P$OR_lo, P$OR_hi)`) ma il suo intervallo di credibilità al 95% include
ancora l'assenza di effetto; la probabilità a posteriori di *una qualsiasi* riduzione è
`r sprintf("%.0f", 100*P$P_OR_lt_1)`%, e scende al `r sprintf("%.0f", 100*P$P_OR_lt_080)`%
per una riduzione clinicamente rilevante (OR < 0,80). Questi risultati distinguono un
plausibile *minor rischio associato* da una *terapia protettiva* consolidata.

**Cosa indica l'adattamento del modello.** Il modello a effetti casuali descrive bene gli
studi, ma il suo esito più informativo è la dispersione tra studi, non la media.
L'eterogeneità è sostanziale ($\tau$ = `r sprintf("%.2f", P$tau)`, I² = `r i2_txt`), per cui
l'OR aggregato riassume una *distribuzione* di effetti specifici per studio, non un unico
valore comune: l'intervallo predittivo per un nuovo studio (OR
`r sprintf("%.2f\u2013%.2f", P$pred_lo, P$pred_hi)`) va da una protezione rilevante a un
modesto danno. Il partial pooling visibile nella Figura 1 attrae le stime imprecise verso
quella distribuzione, ed è per questo che l'intervallo aggregato è più ampio di quanto
suggerirebbe un'analisi a effetti fissi. La posteriori numerica esatta è stata riprodotta
da Hamiltonian Monte Carlo indipendente (R-hat ≈ 1,00, dimensioni campionarie efficaci
nell'ordine delle migliaia, `r n_div` transizioni divergenti): questa ampiezza riflette
fedelmente le evidenze e non un artefatto del calcolo o della scelta del prior.

**Quanto è robusto il risultato?** La direzione dell'effetto è stabile in tutte le
verifiche pre-specificate: nessun singolo studio la determina (leave-one-out OR
`r sprintf("%.2f", min(loo$OR))`–`r sprintf("%.2f", max(loo$OR))`), persiste escludendo le
tre conversioni RR/HR/IRR→OR (OR
`r fmt_ci(sens_prim_or$OR, sens_prim_or$OR_lo, sens_prim_or$OR_hi)`) e limitandosi agli
studi a basso rischio di bias, e non vi è asimmetria del funnel (Egger p =
`r sprintf("%.2f", egg$pval)`). È inoltre stabile tra i prior dominati dai dati (primario,
vago, scettico). L'unica assunzione che rende il risultato decisamente protettivo è il
prior informato dalla letteratura centrato sull'OR 0,62 di Tariq; la verifica formale del
conflitto prior-dati rileva solo una lieve tensione (z = `r sprintf("%.2f", pdc_z)`, p =
`r sprintf("%.2f", pdc_p)`), quindi tale spostamento riflette la ristrettezza del prior e
non un'incompatibilità con i dati aggiornati. I segnali interni più marcati — il
sottoinsieme della sola doxiciclina (OR `r fmt_ci(D$OR, D$OR_lo, D$OR_hi)`) e i contesti
ospedalieri — sono coerenti con un effetto reale ma piccolo, concentrato in agenti e
contesti specifici, benché ciascun intervallo tocchi ancora 1.
:::

## [What this review adds]{.en}[Cosa aggiunge questa revisione]{.it}

::: {.en}
To our knowledge this is the first Bayesian re-analysis of the tetracycline–CDI question
and the first update of Tariq et al. (2018) with the subsequent literature: 20 studies
contribute a curated effect size (11 vs other antibiotics, 9 vs no antibiotic) drawn from
35 included reports, roughly doubling the antibiotic-comparator evidence base.

Three features distinguish it from the earlier meta-analysis. First, it **separates the two
comparator questions** that Tariq pooled — tetracyclines *vs other antibiotics* (the
protocol's PECO) and *vs no antibiotic* — a distinction that materially changes the reading,
because the apparent protection largely tracks the second, confounding-prone comparison.
Second, it reports the **full posterior** — the probabilities P(OR<1) and P(OR<0.80) and a
predictive interval — rather than a single point estimate and confidence interval, making
the residual uncertainty and heterogeneity explicit. Third, it adds a **pre-specified
robustness programme** (prior sensitivity, a formal prior-data conflict test, leave-one-out,
small-study diagnostics, and care-context/design/adjustment meta-regressions) and
cross-validates the exact posterior with MCMC.

Substantively, the headline is a **tempering** of the earlier literature. Tariq's OR ≈ 0.62
reproduces exactly in our hands (`r v_tariq_dl`), but adding the newer, larger
administrative-cohort evidence moves the pooled estimate toward the null (`r v_bridge` for
the combined set), and the confidently protective reading now survives only under a prior
anchored on the older result. The earlier "low risk" conclusion therefore appears driven
partly by comparator mixing and a smaller, earlier evidence base.
:::
::: {.it}
A nostra conoscenza questa è la prima ri-analisi bayesiana della questione tetracicline–ICD
e il primo aggiornamento di Tariq et al. (2018) con la letteratura successiva: 20 studi
contribuiscono con una stima d'effetto curata (11 vs altri antibiotici, 9 vs nessun
antibiotico) tratti da 35 report inclusi, raddoppiando all'incirca la base di evidenze con
comparatore antibiotico.

Tre elementi la distinguono dalla meta-analisi precedente. Primo, **separa le due domande
sul comparatore** che Tariq aveva aggregato — tetracicline *vs altri antibiotici* (il PECO
del protocollo) e *vs nessun antibiotico* — una distinzione che cambia sostanzialmente la
lettura, perché la protezione apparente segue in larga parte il secondo confronto, soggetto
a confondimento. Secondo, riporta la **posteriori completa** — le probabilità P(OR<1) e
P(OR<0,80) e un intervallo predittivo — anziché una singola stima puntuale con intervallo di
confidenza, rendendo esplicite l'incertezza residua e l'eterogeneità. Terzo, aggiunge un
**programma di robustezza pre-specificato** (sensibilità ai prior, test formale di conflitto
prior-dati, leave-one-out, diagnostica dei piccoli studi e meta-regressioni per contesto,
disegno e aggiustamento) e convalida la posteriori esatta con MCMC.

Sul piano sostanziale, il messaggio è un **ridimensionamento** della letteratura
precedente. L'OR ≈ 0,62 di Tariq si riproduce esattamente (`r v_tariq_dl`), ma l'aggiunta
delle evidenze più recenti da grandi coorti amministrative sposta la stima aggregata verso
il valore nullo (`r v_bridge` per l'insieme combinato), e la lettura decisamente protettiva
sopravvive ora solo con un prior ancorato al risultato precedente. La precedente conclusione
di "basso rischio" appare quindi guidata in parte dalla mescolanza dei comparatori e da una
base di evidenze più piccola e datata.
:::

## [Strengths and limitations]{.en}[Punti di forza e limiti]{.it}

::: {.en}
**Strengths.** The review follows a pre-registered PROSPERO protocol and separates
comparators in line with its PECO. The Bayesian random-effects framework propagates the
uncertainty in between-study heterogeneity that frequentist point estimates tend to
understate, and expresses results as directly interpretable probabilities and a predictive
interval. The exact posterior is reproduced by independent HMC sampling, and a broad,
pre-specified sensitivity and diagnostic suite — prior sensitivity, prior-data conflict,
leave-one-out, Egger's test, and three meta-regressions — is reported transparently with
fully reproducible code.

**Limitations.** All evidence is observational, so confounding by indication cannot be
excluded — most acutely in the no-antibiotic comparison. Screening and risk-of-bias
assessment are provisional single-reviewer judgements pending the protocol's duplicate
full-text review; the extraction covariate matrix is not yet populated; and the curated
estimates are still flagged unverified. The "other antibiotics" comparator is a
heterogeneous mix rather than a single clinically matched drug, and three estimates enter as
RR/HR/IRR converted to OR under a rare-outcome approximation. Per-group counts are small, so
the subgroup and meta-regression results are exploratory. Finally, funnel-based diagnostics
have limited power at k = `r nrow(prim_abx)`, so publication bias cannot be excluded despite
the absence of detectable asymmetry, and neither individual-participant data nor head-to-head
comparisons against specific agents were available.
:::
::: {.it}
**Punti di forza.** La revisione segue un protocollo PROSPERO pre-registrato e separa i
comparatori in linea con il suo PECO. Il modello bayesiano a effetti casuali propaga
l'incertezza sull'eterogeneità tra studi che le stime puntuali frequentiste tendono a
sottostimare, ed esprime i risultati come probabilità direttamente interpretabili e un
intervallo predittivo. La posteriori esatta è riprodotta da campionamento HMC indipendente,
e un ampio insieme pre-specificato di analisi di sensibilità e diagnostica — sensibilità ai
prior, conflitto prior-dati, leave-one-out, test di Egger e tre meta-regressioni — è
riportato in modo trasparente con codice pienamente riproducibile.

**Limiti.** Tutte le evidenze sono osservazionali, per cui il confondimento da indicazione
non può essere escluso — in modo più acuto nel confronto vs nessun antibiotico. Lo screening
e la valutazione del rischio di bias sono giudizi provvisori di un singolo revisore, in
attesa della doppia revisione full-text prevista dal protocollo; la matrice delle covariate
di estrazione non è ancora popolata; e le stime curate risultano tuttora non verificate. Il
comparatore "altri antibiotici" è un insieme eterogeneo e non un singolo farmaco clinicamente
confrontabile, e tre stime entrano come RR/HR/IRR convertiti in OR con un'approssimazione per
esiti rari. Le numerosità per gruppo sono piccole, quindi i risultati dei sottogruppi e delle
meta-regressioni sono esplorativi. Infine, la diagnostica basata sul funnel ha potenza
limitata con k = `r nrow(prim_abx)`, per cui il bias di pubblicazione non può essere escluso
nonostante l'assenza di asimmetria rilevabile, e non erano disponibili né dati individuali
dei partecipanti né confronti diretti con agenti specifici.
:::

## [Discussion]{.en}[Discussione]{.it}

::: {.en}
**From evidence to practice.** The Interpretation above summarises the statistical picture —
a pooled OR below 1 that remains compatible with no effect, dominated by between-study
heterogeneity, and confidently protective only under a literature-anchored prior. What does
that support in practice?

**Is tetracycline preferable to macrolides or other agents to reduce CDI?** Not on this
evidence. First, no included study provides a head-to-head tetracycline-versus-macrolide
comparison; the comparator is a heterogeneous mix of "other antibiotics", so a
macrolide-specific preference would be an extrapolation. Second, the pooled effect is
uncertain (credible interval crossing 1, substantial heterogeneity). Third, the protective
interpretation is fragile, depending on prior belief. The defensible statement is narrower:
tetracyclines — particularly doxycycline — are biologically plausible lower-CDI-risk agents
and are unlikely to be worse than typical comparators, with a *possible* modest advantage.
That can reasonably inform stewardship *when the spectrum required is genuinely equivalent*,
but it does not meet the bar for a formal recommendation to prefer them, and certainly not
specifically over macrolides.

**Implications for future studies.**

- Use active-comparator, new-user cohort designs (or target-trial emulation) with an
  explicit, clinically matched comparator — e.g. doxycycline vs a macrolide for
  community-acquired pneumonia — rather than pooling against undifferentiated "other antibiotics".
- Reduce confounding by indication: adjust for or match on prior CDI, PPI use, hospitalisation,
  age and comorbidity; treat case-control evidence cautiously given its design-related divergence.
- Standardise CDI ascertainment (testing protocols, exposure and outcome windows) so
  heterogeneity reflects biology rather than measurement.
- Analyse by agent (doxycycline vs minocycline vs tigecycline) to isolate the doxycycline signal.
- Pre-register and report completely to limit publication bias, and consider
  individual-participant-data meta-analysis to resolve the heterogeneity the predictive interval exposes.
- Where clinically and ethically appropriate, a pragmatic randomised trial with CDI as a
  pre-specified secondary outcome would provide the confirmatory evidence observational data cannot.
:::
::: {.it}
**Dalle evidenze alla pratica.** L'Interpretazione qui sopra riassume il quadro statistico —
un OR aggregato inferiore a 1 che resta compatibile con l'assenza di effetto, dominato
dall'eterogeneità tra studi e decisamente protettivo solo con un prior ancorato alla
letteratura. Che cosa supporta tutto ciò nella pratica?

**Le tetracicline sono preferibili ai macrolidi o ad altri agenti per ridurre l'ICD?** Non
sulla base di queste evidenze. Primo, nessuno studio incluso fornisce un confronto diretto
tetraciclina-vs-macrolide; il comparatore è un insieme eterogeneo di "altri antibiotici",
perciò una preferenza specifica sui macrolidi sarebbe un'estrapolazione. Secondo, l'effetto
aggregato è incerto (intervallo di credibilità che attraversa 1, eterogeneità sostanziale).
Terzo, l'interpretazione protettiva è fragile e dipende dalla convinzione a priori.
L'affermazione difendibile è più circoscritta: le tetracicline — in particolare la
doxiciclina — sono agenti biologicamente plausibili a minor rischio di ICD ed è improbabile
che siano peggiori dei comparatori abituali, con un *possibile* modesto vantaggio. Ciò può
ragionevolmente orientare la stewardship *quando lo spettro richiesto è realmente
equivalente*, ma non raggiunge la soglia per una raccomandazione formale a preferirle, e
certamente non specificamente rispetto ai macrolidi.

**Implicazioni per gli studi futuri.**

- Usare disegni di coorte active-comparator, new-user (o emulazione di target trial) con un
  comparatore esplicito e clinicamente confrontabile — es. doxiciclina vs un macrolide nella
  polmonite acquisita in comunità — invece di aggregare contro "altri antibiotici" indistinti.
- Ridurre il confondimento da indicazione: aggiustare o appaiare per ICD pregressa, uso di
  IPP, ospedalizzazione, età e comorbidità; trattare con cautela le evidenze caso-controllo
  data la divergenza legata al disegno.
- Standardizzare l'accertamento dell'ICD (protocolli di test, finestre di esposizione ed
  esito) affinché l'eterogeneità rifletta la biologia e non la misurazione.
- Analizzare per agente (doxiciclina vs minociclina vs tigeciclina) per isolare il segnale
  della doxiciclina.
- Pre-registrare e riportare in modo completo per limitare il publication bias, e valutare
  una meta-analisi su dati individuali dei partecipanti per risolvere l'eterogeneità che
  l'intervallo predittivo evidenzia.
- Ove clinicamente ed eticamente appropriato, uno studio randomizzato pragmatico con l'ICD
  come esito secondario pre-specificato fornirebbe la conferma che i dati osservazionali non
  possono dare.
:::

::: {.callout-warning}
::: {.en}
**Caveats.** All included studies are observational; Newcastle-Ottawa scores are
predominantly in the low-risk range, but the assessment is provisional pending duplicate
full-text review. Estimates convert RR/HR to OR under a rare-outcome approximation. This
is evidence synthesis for research and education, not treatment guidance for individual
patients.
:::
::: {.it}
**Avvertenze.** Tutti gli studi inclusi sono osservazionali; i punteggi Newcastle-Ottawa
sono prevalentemente nell'intervallo di basso rischio, ma la valutazione è provvisoria in
attesa della doppia revisione del full-text. Le stime convertono RR/HR in OR con
un'approssimazione per esiti rari. Si tratta di una sintesi delle evidenze a scopo di
ricerca ed educativo, non di una guida terapeutica per il singolo paziente.
:::
:::

## [References]{.en}[Bibliografia]{.it}

::: {.en}
Bibliographic entries for the studies in the synthesis are cited in Table 1 and listed below.
Author and year are taken from the extraction; titles, journals and DOIs in `references.bib`
are a scaffold to be verified/completed by the author (Tariq 2018 and a few studies whose DOIs
came from the source files are complete).
:::
::: {.it}
Le voci bibliografiche degli studi nella sintesi sono citate nella Tabella 1 ed elencate qui
sotto. Autore e anno provengono dall'estrazione; titoli, riviste e DOI in `references.bib` sono
uno scheletro da verificare/completare (Tariq 2018 e alcuni studi con DOI dai file sorgente
sono completi).
:::

::: {#refs}
:::

---

::: {.cap}
::: {.en}
Reproducible source: `report.qmd`. Bayesian normal-normal hierarchical model (`bayesmeta`, exact numerical posterior), confirmed by `brms`/Stan MCMC. Re-render with `quarto render report.qmd`.
:::
::: {.it}
Sorgente riproducibile: `report.qmd`. Modello gerarchico normale-normale bayesiano (`bayesmeta`, posteriori numerica esatta), confermato con MCMC `brms`/Stan. Ri-generare con `quarto render report.qmd`.
:::
:::
